---
title: "analysis"
author: "Christopher Odoom"
date: "2025-11-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
dir.create("figures", showWarnings = FALSE)
dir.create("outputs", showWarnings = FALSE)
```

## Add required packages 

```{r package import, echo=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(lme4)
library(readr)
library(scales)
library(patchwork)
library(rlang)
library(purrr)
library(survey)

```

## Load the datasets

```{r Load sedentary and sleep data, echo=FALSE, warning=FALSE}

## Load near sedentary (no sleep data) time data
sedentary_hip_df <- read_csv('data/nosleep_data.csv.gz')

sedentary_wrist_df <- read_csv('data/wrist_df.csv.gz')

## Load sleep data
sleep_df <- read_csv('data/sleep_est_data.csv.gz')


```

## Inspect the loaded dataset

```{r Na checker}
## function to check nas
rows_with_na <- function(df, column) {
  df[is.na(df[[column]]), ]
}

head(sleep_df)
```



```{r}
sleep_df <- sleep_df %>% select(ID,Day, Hour, percent_sleep_nonwear)
head(sleep_df)
```


```{r}
## Check number of rows and columns in the wrist datasets

dim(sedentary_wrist_df)
dim(sedentary_hip_df)
dim(sleep_df)

## Check the number of unique IDs in each dataset
nidsw <- length(unique(sedentary_wrist_df$ID))
nidsd <- length(unique(sedentary_hip_df$ID))
nids <- length(unique(sleep_df$ID))
print(paste0("The number of unique ID's in the wrist dataset",' ',  nidsw))
print(paste0("The number of unique ID's in the hip dataset",' ',  nidsd))
print(paste0("The number of unique ID's in the sleep dataset",' ',  nids))


## Find the list of unique IDs in the hip dataset that are not in the wrist dataset and write a txt file
ids_in_hip_not_wrist <- setdiff(unique(sedentary_hip_df$ID), unique(sedentary_wrist_df$ID))
writeLines(as.character(ids_in_hip_not_wrist), "outputs/ids_in_hip_not_wrist.txt")

```
```{r}
length(ids_in_hip_not_wrist)

```




```{r}
head(sedentary_hip_df)

## rewrist the text file by adding the Datset column to the text file from the hip data
## unique IDs in the hip dataset that are not in the wrist dataset with their corresponding Dataset value from the hip dataset

ids_in_hip_not_wrist_df <- sedentary_hip_df %>% 
  filter(ID %in% ids_in_hip_not_wrist) %>% 
  select(ID, Dataset) %>% 
  distinct()
## write a txt file with the ID and Dataset columns

writeLines(paste(ids_in_hip_not_wrist_df$Dataset,ids_in_hip_not_wrist_df$ID, sep = ","), "outputs/ids_in_hip_not_wrist.txt")



```

```{r Inspect the data set, echo=FALSE, warning=FALSE}
## See some of the sedentary data from hip model 

head(sedentary_hip_df)
## Remove one column
colnames(sedentary_hip_df)[colnames(sedentary_hip_df)=='PercentSedentary'] <- 'percent_sedentary_hip'

colnames(sedentary_hip_df)[colnames(sedentary_hip_df)=='Date'] <- 'Day'

nidsd<- length(unique(sedentary_hip_df$ID))
print(paste0("The number of unique ID's in the sedentary data",' ',  nidsd))

head(sedentary_hip_df)

sedentary_h.df <- sedentary_hip_df %>% filter(PartialDay!='Yes')
## check for Nas
rows_with_na(sedentary_h.df, "percent_sedentary_hip")

head(sedentary_h.df)
```



```{r}
## See some of the sedentary data from wrist model
head(sedentary_wrist_df)

## Rename some columns for easy interprability
colnames(sedentary_wrist_df)[colnames(sedentary_wrist_df)=='percent_sitting'] <- 'percent_sedentary_wrist'
colnames(sedentary_wrist_df)[colnames(sedentary_wrist_df)=='Date'] <- 'Day'
nidsw<- length(unique(sedentary_wrist_df$ID))
print(paste0("The number of unique ID's in the wrist dataset",' ',  nidsw))

head(sedentary_wrist_df)
```


```{r keep only required columns for the wrist dataset, echo=FALSE, warning=FALSE}

head(sedentary_wrist_df)

sedentary_w.df = sedentary_wrist_df %>% select(ID, Day, Hour, percent_sedentary_wrist)

head(sedentary_w.df)
```



### Combining weights from 2 survey circles
### According to the NHANES analytic guidlines, we must verify the following

1. Be aware of sample design changes between cycles.
2. Use the file documentation to verify that data items collected in the cycles that are to be combined are comparable in question wording, methods, and inclusion or exclusion criteria.
3. Examine the validity of the assumption that no trend in the estimate exists over the period being combined.

We can test 3 by answering the following questions
- Do average sedentary time increase or decrease between 2011-2012 and 2013-2014?
- Do average sleep change between cycles?

Since we are combining 2 cycles to get a 4 year cylce, we need to recalculate the weights as
   $ \textbf{Exam_weights_recal} = \text{Exam_weight}\times \frac{2}{4}$


```{r calculate combined weights based on HNANES guidlines}
## Recalculate examination weights(for combined cycle)

sedentary_h.df$Exam_weight_recal <- 0.5*sedentary_h.df$Exam_weight

head(sedentary_h.df)
length(unique(sedentary_h.df$ID))
```

```{r}
## Use dplyr to perform merge on (ID, Day,Hour)
sedentary.df <- dplyr::inner_join(y=sedentary_h.df,x=sedentary_w.df, by=c('ID', 'Day', 'Hour'))

head(sedentary.df)
dim(sedentary.df)
## number of unique IDs in the merged dataset
nidss<- length(unique(sedentary.df$ID))
print(paste0("The number of unique ID's in the merged sedentary dataset",' ',  nidss))
```



```{r Inspect the sleep data set, echo=FALSE, warning=FALSE}
## See some of the sleep data

head(sleep_df)

## rename some columns for easy interprability
colnames(sleep_df)[colnames(sleep_df) =='percent_sleep_nonwear'] <- 'percent_sleep'
colnames(sleep_df)[colnames(sleep_df) =='percent_wear'] <- 'percent_wake'
head(sleep_df)
nids<- length(unique(sleep_df$ID))
print(paste0("The number of unique ID's",' ',  nids))
```

The total participants for the two cycles is 14693 but 159 participants either have no complete day data or we failed to process the sleep data (89 participants are due to the later). 
There study uses a total of 14534 participants.

```{r check missing sleep row}
rows_with_na <- function(df, column) {
  df[is.na(df[[column]]), ]
}

rows_with_na(sleep_df, "percent_sleep")

sleep_df <- sleep_df[complete.cases(sleep_df), ]

head(sleep_df)
nids<- length(unique(sleep_df$ID))
print(paste0("The number of unique ID's",' ',  nids))
```

```{r}
# # IDs in sed_all but missing from sleep_all
# setdiff(unique(sed_all$ID), unique(sleep_all$ID))
# 
# # IDs in sleep_all but missing from sed_all
# setdiff(unique(sleep_all$ID), unique(sed_all$ID))
# 
# # Inspect their values
# sleep_sed.df %>% filter(ID %in% setdiff(unique(sleep_all$ID), unique(sed_all$ID))) %>%
#   select(ID, percent_sleep, percent_sedentary)
```


## Get a merge data set of the two dataset (Also keep few variable: important)


```{r merge the dataset, echo=FALSE, warning=FALSE}
## Use dplyr to perform merge on (ID, Day,Hour)
sleep_sed.df <- dplyr::inner_join(y=sedentary.df,x=sleep_df, by=c('ID', 'Day', 'Hour'))

head(sleep_sed.df)
nidslsd<- length(unique(sleep_df$ID))
print(paste0("The number of unique ID's in merged data",' ',  nidslsd))

## check for Nas
rows_with_na(sleep_sed.df, "percent_sedentary_wrist")


length(unique(sleep_df$ID))
length(unique(sleep_sed.df$ID))
```



```{r merge sanity check, echo=FALSE, warning=FALSE}
(table(sleep_sed.df$PartialDay))

# check nas
#summary(sleep_sed.df)
#sleep_sed.df[rowSums(is.na(sleep_sed.df)) > 0, ]


```

We have a successful merge with clean data no partial days and no missing sleep or sedentary row. 


## Select required columns for analysis (Make a copy of the original merged data)

```{r select columns, echo=FALSE, warning=FALSE}

# sl_sed.df <- sleep_sed.df %>% select(ID, Day,DayType, Hour,percent_sedentary,
#                                      percent_sleep, Age,Gender,
#                                      Weight, Height,Exam_weight)

sl_sed.df <- sleep_sed.df 
head(sl_sed.df)
#sl_sed.df[rowSums(is.na(sl_sed.df)) > 0, ]
colSums(is.na(sl_sed.df))
#sleep_sed.df <- sleep_sed.df[!is.na(sleep_sed.df$percent_sedentary), ]



rows_with_na(sl_sed.df, "percent_sedentary_wrist")

```

## Examine if there is trend in sleep/sedentary times between the two cycles
We will use cycle specific examination weights, PSU and Stratum to get estimates

When you combine multiple NHANES cycles into one pooled estimate (2011–2014), CDC implicitly assumes:
The true population mean for your outcome did not materially change between 2011–2012 and 2013–2014.



```{r}
## Define cycles
sl_sed.df$cycle <- ifelse(sl_sed.df$Dataset == "2011-12", 1, 2)

#Create two cycle-specific survey design objects
library(survey)

# 2011–2012 cycle
des_1112 <- svydesign(
  id     = ~PSU,
  strata = ~Stratum,
  weights = ~Exam_weight,
  nest   = TRUE,
  data   = subset(sl_sed.df, cycle == 1)
)

# 2013–2014 cycle
des_1314 <- svydesign(
  id     = ~PSU,
  strata = ~Stratum,
  weights = ~Exam_weight,
  nest   = TRUE,
  data   = subset(sl_sed.df, cycle == 2)
)

```



##  Compute survey weighted means for the two cycles

```{r}
# 1) Percent sleep by cycle
sleep_1112 <- svymean(~percent_sleep, des_1112, na.rm = TRUE)
sleep_1314 <- svymean(~percent_sleep, des_1314, na.rm = TRUE)

# 2) Percent sedentary by cycle
sed_1112 <- svymean(~percent_sedentary_wrist, des_1112, na.rm = TRUE)
sed_1314 <- svymean(~percent_sedentary_wrist, des_1314, na.rm = TRUE)

```



```{r}
sleep_1112
sleep_1314
sed_1112
sed_1314

```

## Trend testing using combined design

Because we’re not pooling for estimates, we leave weights unchanged and do not divide by 2. We use cycle-specific Exam_weight.


```{r}
des_comb <- svydesign(
  id     = ~PSU,
  strata = ~Stratum,
  weights = ~Exam_weight,
  nest   = TRUE,
  data   = sl_sed.df
)

# numeric cycle indicator (1 = 2011–12, 2 = 2013–14)
sl_sed.df$cycle_num <- ifelse(sl_sed.df$Dataset == "2011-12", 0, 1)
des_comb <- update(des_comb, cycle_num = sl_sed.df$cycle_num)

# Trend test for sleep
trend_sleep <- svyglm(percent_sleep ~ cycle_num, design = des_comb)
summary(trend_sleep)

# Trend test for sedentary
trend_sed <- svyglm(percent_sedentary_wrist ~ cycle_num, design = des_comb)
summary(trend_sed)

```

```{r}
## 95 % CI
confint(trend_sed)
```


```{r Table of estimates}


cycle_summary <- svyby(
  ~percent_sleep + percent_sedentary_wrist,
  ~Dataset,
  design = des_comb,
  svymean,
  vartype = c("se", "ci"),
  na.rm = TRUE
)

cycle_summary

```

To examine whether a temporal trend existed between the two cycles (2011–2012 and 2013–2014), we computed cycle-specific survey-weighted estimates using cycle-specific examination weights, masked variance units (PSU, Stratum), and domain analysis. We then tested for linear trend by fitting a survey-weighted regression model with cycle as a numeric predictor. If no significant temporal trend was detected, the assumption of stability was considered satisfied and pooling into a 4-year analytic period (2011–2014) was justified. If a significant trend was found, cycle-specific estimates were reported.

## Results

Sleep/nonwear time increased slightly from 48.9% (SE = 0.36) in 2011–2012 to 50.3% (SE = 0.33) in 2013–2014 (β = 1.43, p = 0.0065). In contrast, sedentary time remained stable across cycles: 75.75% (SE = 0.22) in 2011–2012 vs 76.34% (SE = 0.19) in 2013–2014 (β = 0.38, p = 0.323). Because percent sedentary time—the primary outcome—showed no significant temporal trend, pooling 2011–2014 cycles was considered appropriate for subsequent analyses.

Look into this properly


## Pooling 2011–2014 cycles for analysis

## Ploting percent sendentary and sleep

```{r survey weighted averages: Function, echo=FALSE, warning=FALSE}


plot_hourly_average_weighted <- function(df, metric = "percent_sedentary_wrist") {

  # 1. Survey design using pooled 2011–2014 exam weights
  des <- svydesign(
    id     = ~PSU,
    strata = ~Stratum,
    weights = ~Exam_weight_recal,
    nest   = TRUE,
    data   = df
  )

  # 2. Compute weighted hourly estimates
  hourly_avg <- svyby(
    as.formula(paste0("~", metric)),
    ~Hour,
    design = des,
    svymean,
    na.rm = TRUE
  ) %>%
    rename(avg_value = !!sym(metric))

  # 3. Plot
  p <- hourly_avg %>%
    ggplot(aes(x = Hour, y = avg_value)) +
    geom_line(size = 1.2, color = "violet") +
    geom_point(size = 2.5, color = "grey30") +
    scale_y_continuous(labels = percent_format(scale = 1), limits = c(10, 100)) +
    scale_x_continuous(breaks = seq(1, 24, 2)) +
    labs(
      x = "Hour of Day",
      y = paste0("% ", gsub("_", " ", gsub("^percent_?", "", metric)))
    ) +
    theme_classic(base_size = 14)

  return(p)
}


```



## Produce Example plots of the overall hourly average percent sedentary and sleep time across the day (using pooled 2011–2014 weights)

```{r sedentary/sleep time, echo=FALSE, warning=FALSE}
p1 <- plot_hourly_average_weighted(sl_sed.df, "percent_sedentary_wrist")
p2 <- plot_hourly_average_weighted(sl_sed.df, "percent_sedentary_hip")

p3 <- plot_hourly_average_weighted(sl_sed.df, "percent_sleep")
p1+p2+p3
```




## Percent sleep and near sedentary stratified by other variables
### no sleep accounted for in the sed

```{r define function to plot, echo=FALSE, warning=FALSE}


plot_hourly_stratified_weighted <- function(df, 
                                            metric = "percent_sedentary_wrist", 
                                            group_var = "DayType") {

  # Convert group_var to symbol
  group_sym <- sym(group_var)

  # 1. Survey design using pooled 2011–2014 NHANES weights
  des <- svydesign(
    id     = ~PSU,
    strata = ~Stratum,
    weights = ~Exam_weight_recal,
    nest   = TRUE,
    data   = df
  )

  # 2. Weighted hourly mean by group using svyby()
  hourly_avg <- svyby(
    as.formula(paste0("~", metric)),
    by = as.formula(paste("~", group_var, "+ Hour")),
    design = des,
    FUN = svymean,
    na.rm = TRUE
  ) %>%
    rename(avg_value = !!sym(metric))

  # 3. Plot results
  p <- hourly_avg %>%
    ggplot(aes(x = Hour, y = avg_value, color = !!group_sym)) +
    geom_line(size = 1) +
    geom_point(size = 2) +
    scale_y_continuous(labels = percent_format(scale = 1), limits = c(10, 100)) +
    scale_x_continuous(breaks = seq(1, 24, 2)) +
    labs(
      x = "Hour of Day",
      y = paste0("% ", gsub("_", " ", gsub("^percent_?", "", metric))),
      color = group_var
    ) +
    theme_classic(base_size = 14)

  return(p)
}

## This adds 95% CI to the plots

plot_hourly_stratified_weighted_CI <- function(df, 
                                               metric = "percent_sedentary", 
                                               group_var = "DayType") {

  group_sym <- sym(group_var)

  # 1. Survey design using 2011–2014 pooled weight
  des <- svydesign(
    id     = ~PSU,
    strata = ~Stratum,
    weights = ~Exam_weight_recal,
    nest   = TRUE,
    data   = df
  )

  # 2. Weighted hourly means + CIs by group
  hourly_avg <- svyby(
    as.formula(paste0("~", metric)),
    by = as.formula(paste("~", group_var, "+ Hour")),
    design = des,
    FUN = svymean,
    na.rm = TRUE,
    vartype = c("ci")  # gives ci_l and ci_u
  ) %>% 
    rename(avg_value = !!sym(metric))

  # 3. Plot
  p <- hourly_avg %>%
    ggplot(aes(x = Hour, y = avg_value, color = !!group_sym, fill = !!group_sym)) +
    
    # CI ribbon
    geom_ribbon(aes(ymin = ci_l, ymax = ci_u), alpha = 0.7, color = NA) +
    
    # Lines
    geom_line(size = 1.1) +
    
    # Points
    geom_point(size = 2) +

    # Formatting
    scale_y_continuous(labels = percent_format(scale = 1), limits = c(10, 100)) +
    scale_x_continuous(breaks = seq(1,24,2)) +
    labs(
      x = "Hour of Day",
      y = paste0("% ", gsub("_", " ", gsub("^percent_?", "", metric))),
      color = group_var,
      fill  = group_var
    ) +
    theme_classic(base_size = 14)

  return(p)
}


```




```{r cycles mean curves}
#head(sl_sed.df)
p3 <- plot_hourly_stratified_weighted_CI(sl_sed.df, metric="percent_sedentary_wrist", group_var="Dataset")
p4 <- plot_hourly_stratified_weighted_CI(sl_sed.df, metric="percent_sleep", group_var="Dataset")


(p3 + p4) +
  plot_layout(
    guides = "collect"    # one shared legend
  ) &
  theme(
    legend.position = "right"   # or "bottom", "top", etc.
  )
```




```{r, sleep and sedentary plot stratified by weekday without CI}
p5 <- plot_hourly_stratified_weighted_CI(sl_sed.df, metric="percent_sedentary_wrist", group_var="DayType")

p6 <- plot_hourly_stratified_weighted_CI(sl_sed.df, metric="percent_sedentary_hip", group_var="DayType")

p7 <- plot_hourly_stratified_weighted_CI(sl_sed.df, metric="percent_sleep", group_var="DayType")


(p5 + p6+p7) +
  plot_layout(
    guides = "collect"    # one shared legend
  ) &
  theme(
    legend.position = "right"   # or "bottom", "top", etc.
  )
```

## Check available participant per cycle

```{r, check the number of unique ID dataset}
sl_sed.df %>% group_by(Dataset) %>% summarise(n_IDs = n_distinct(ID))
```


## Table 1: Data distribution by cycle

```{r}
library(dplyr)
library(tidyr)

# Count number of valid days per ID
df_days <- sl_sed.df %>%
  distinct(ID, Day, Dataset) %>%
  group_by(ID, Dataset) %>%
  summarise(n_days = n(), .groups = "drop")

# Distribution by dataset
table_by_cycle <- df_days %>%
  count(Dataset, n_days, name = "n") %>%
  group_by(Dataset) %>%
  mutate(
    percent = 100 * n / sum(n),
    value = paste0(n, " (", round(percent, 1), ")")
  ) %>%
  ungroup() %>%
  select(n_days, Dataset, value) %>%
  pivot_wider(names_from = Dataset, values_from = value)

# Total distribution
table_total <- df_days %>%
  count(n_days, name = "n") %>%
  mutate(
    percent = 100 * n / sum(n),
    `Total (2011–14)` = paste0(n, " (", round(percent, 1), ")")
  ) %>%
  select(n_days, `Total (2011–14)`)

# Main table
table_final <- table_by_cycle %>%
  left_join(table_total, by = "n_days") %>%
  arrange(n_days) %>%
  mutate(n_days = as.character(n_days))

# Available sample row
available_row <- df_days %>%
  group_by(Dataset) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(value = paste0(n, " (100)")) %>%
  select(Dataset, value) %>%
  pivot_wider(names_from = Dataset, values_from = value) %>%
  mutate(
    n_days = "Available sample (1–7)",
    `Total (2011–14)` = paste0(nrow(df_days), " (100)")
  )

# Total row
total_row <- available_row %>%
  mutate(n_days = "Total")

# Combine
table_final <- bind_rows(
  table_final,
  available_row,
  total_row
)

# Print
knitr::kable(
  table_final,
  col.names = c("Full days", "2011–2012", "2013–2014", "Total (2011–14)")
)
```



```{r}
knitr::kable(
  table_final,
  col.names = c("Full days", "2011–2012", "2013–2014", "Total (2011–14)")
)
```



## Next: analysis of Sedentary time in actual wake time

## Summarise sedentary time taking into account sleep


```{r define the function to summarise sedentary and sleep time for different cutoffs, echo=FALSE, warning=FALSE}

## A more robust function to calculate the same summary but also ensures that all IDs are included in the summary even if they have no hours classified as sleep or sedentary for a given cutoff. This way we can track how the number of hours included in each category changes with different cutoffs.

compute_sleep_sed_stats_fixed <- function(df, cutoffs) {

  all_IDs <- unique(df$ID)

  results_sed <- map(cutoffs, function(cut) {
    ## Show progress
    print(paste0("Processing cutoff: ", cut, "% sleep"))
    
    # ----- 1) Sedentary (wake) summary -----
    sed_df <- df %>%
      filter(percent_sleep < cut) %>%
      group_by(ID, Day) %>%
      summarise(
        total_sed_hours = sum(percent_sedentary_wrist / 100, na.rm = TRUE),
        n_hours = n(),
        sleep_hrs =24 - n_hours, # Assuming 24 hours in a day, sleep hours = 24 - wake hours
        sd_sed = sd(percent_sedentary_wrist / 100, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      # Ensure all IDs are included
      right_join(distinct(df, ID, Day), by = c("ID", "Day")) %>%
      mutate(
        total_sed_hours = replace_na(total_sed_hours, 0),
        n_hours = replace_na(n_hours, 0),
        sleep_hrs = replace_na(sleep_hrs, 24), # If no wake hours, assume all 24 hours are sleep
        sd_sed = replace_na(sd_sed, 0),
        cutoff = cut
      )
list(sed_df = sed_df)
  })
  
# results_sleep <- map(cutoffs, function(cut) {
#     ## Show progress
#     print(paste0("Processing cutoff: ", cut, "% sleep"))
#     
#     # ----- 2) Sleep summary -----
#     sleep_df <- df %>%
#       filter(percent_sleep >= cut) %>%
#       group_by(ID, Day) %>%
#       summarise(
#         total_sleep_hours = sum(percent_sleep / 100, na.rm = TRUE),
#         n_hours = n(),
#         wake_hrs =24 - n_hours, # Assuming 24 hours in a day, wake hours = 24 - sleep hours
#         sd_sleep = sd(percent_sleep / 100, na.rm = TRUE),
#         .groups = "drop"
#       ) %>%
#       # Ensure all IDs are included
#       right_join(distinct(df, ID, Day), by = c("ID", "Day")) %>%
#       mutate(
#         total_sleep_hours = replace_na(total_sleep_hours, 0),
#         n_hours = replace_na(n_hours, 0),
#         wake_hrs = replace_na(wake_hrs, 24), # If no sleep hours, assume all 24 hours are wake
#         sd_sleep = replace_na(sd_sleep, 0),
#         cutoff = cut
#       )  
# list(sleep_df = sleep_df)
#   })
#   

  # Combine results across all cutoffs
    sedentary_big <- bind_rows(map(results_sed, "sed_df"))
    #sleep_big     <- bind_rows(map(results_sleep, "sleep_df"))
  
    
  return(
    sed_summ_df = sedentary_big
  )
}



```





## Get the summary of sedentary and sleep time for different cutoffs

```{r, get the data for different cutoffs, echo=FALSE, warning=FALSE}
cutoffs <- seq(25,100,1)

out <- compute_sleep_sed_stats_fixed(sleep_sed.df, cutoffs)

sed_all  <- out


summary(sed_all)
#length(unique(sleep_all$ID))
length(unique(sed_all$ID))
```



## View sedentary time summary

```{r, view summary, echo=FALSE, warning=FALSE}
dim(sed_all)
length(unique(sed_all$ID))

head(sed_all)

## check nas
summary(sed_all)
```

```{r}
## Make a plot of Day on the x and total y1=sedentary hours and y2= number of hours on the y, facet cutoff(select 2 cutoffs, 35 and 95) for participant ID 62161 as an example
sed_all %>% filter(ID == 62161, cutoff %in% c(35, 75,95)) %>%
  ggplot(aes(x = Day)) +
  geom_line(aes(y = total_sed_hours, color = "Sedentary Hours"), size = 1.2) +
  geom_point(aes(y = total_sed_hours, color = "Sedentary Hours"), size = 2.5) +
  geom_line(aes(y = n_hours, color = "Wake Hours"), size = 1.2, linetype = "dashed") +
  geom_point(aes(y = n_hours, color = "Wake Hours"), size = 2.5) +
  labs(
    x = "Day",
    y = "Hours",
    title = "Sedentary Time vs Wake Hours for a Participant",
    color = NULL
  ) +
  scale_color_manual(values = c("Sedentary Hours" = "blue", "Wake Hours" = "black")) +
  theme_classic(base_size = 20) +
  facet_wrap(~cutoff)

```





## plot the summary of sedentary and sleep time for different cutoffs: A function to perform this plotting for both sedentary and sleep time summaries.

```{r, define plotting function, echo=FALSE, warning=FALSE}
plot_cutoff_summary <- function(summary_df, 
                                type = c("sedentary", "sleep")) {
  
  library(dplyr)
  library(ggplot2)
  
  type <- match.arg(type)
  
  # Select correct columns depending on type
  if (type == "sedentary") {
    df_plot <- summary_df %>%
      group_by(cutoff) %>%
      summarise(
        avg_hours = mean(total_sed_hours, na.rm = TRUE),
        avg_n_hours = mean(n_hours, na.rm = TRUE)
      )
    y_label <- "Average Sedentary Hours (per Day)"
  } else {
    df_plot <- summary_df %>%
      group_by(cutoff) %>%
      summarise(
        avg_hours = mean(total_sleep_hours, na.rm = TRUE),
        avg_n_hours = mean(n_hours, na.rm = TRUE)
      )
    y_label <- "Average Sleep Hours (per Day)"
  }
  
  # Plot
  p <- df_plot %>%
    ggplot(aes(x = cutoff)) +
    
    # Main line (sedentary or sleep hours)
    geom_line(aes(y = avg_hours, color = type), size = 1.2) +
    geom_point(aes(y = avg_hours, color = type), size = 2.5) +
    
    # Secondary line (average n hours)
    geom_line(aes(y = avg_n_hours, color = "n_hours"), 
              size = 1.2, linetype = "dashed") +
    geom_point(aes(y = avg_n_hours, color = "n_hours"), size = 2.5) +
    
    labs(
      x = "% Sleep Cutoff",
      y = "Hours",
      #title = paste("Cutoff Sweep for", type, "(Avg Daily Hours Across All Participants)"),
      color = NULL
    ) +
    scale_color_manual(
      values = c(
        "sedentary" = "blue",
        "sleep" = "purple",
        "n_hours" = "black"
      ),
      labels = c(
        "sedentary" = "Avg Sitting Hours",
        "sleep" = "Avg Sleep Hours",
        "n_hours" = "Avg number of Hours Included"
      )
    ) +
    theme_classic(base_size = 14)+theme(legend.position = 'bottom')+scale_x_continuous(
    limits = c(25, 100),
    breaks = seq(25, 100, by = 5)
  )+ scale_y_continuous(
    limits = c(0, 34),
    breaks = seq(0, 24, by = 2)
  )
  
  return(p)
}

```



```{r}
psed <- plot_cutoff_summary(sed_all, type = "sedentary")
#psleep <- plot_cutoff_summary(sleep_all, type = "sleep")
#+psleep
psed
```









## modeling (cutoff vs sleep/ sedentary time)

## Prepare data for modeling


```{r sedentary: Compute per-ID daily averages from the hourly dataset}

tail(sed_all)
person_level_sed.fun <- function(df, cut){
  # df is the summary dataset for sedentary time with columns ID, Day, total_sed_hours, n_hours, cutoff
  
person_level_sedentary <- df %>% filter(cutoff==cut) %>% 
  group_by(ID) %>% 
  summarise(mean_sed_time = mean(total_sed_hours),
            avg_wake_hours = mean(n_hours),
            avg_sleep_hours = 24 - avg_wake_hours, # Assuming 24 hours in a day
            n_days = n(),
            day_sd=sd(total_sed_hours),
            cutoff=cut,
            .groups = "drop"
  ) 
return(person_level_sedentary)
}

# Apply function to all cutoffs and combine
person_level_sed_all <- bind_rows(
  lapply(cutoffs, function(cut) person_level_sed.fun(df = sed_all, cut = cut))
)

# Check
head(person_level_sed_all)
```

## Function to compute person-level sleep summary for each cutoff and combine results across cutoffs

```{r summarise sleep}


# Function for a single cutoff (sleep)
person_level_sleep.fun <- function(df, cut){
  df %>%
    filter(cutoff == cut) %>%
    group_by(ID) %>%
    summarise(
      mean_sleep_time = mean(total_sleep_hours),
      avg_sleep_hours = mean(n_hours),
      n_days_sleep   = n(),
      day_sd_sleep   = sd(total_sleep_hours),
      cutoff=cut,
      .groups = "drop"
    )
}



# Apply function to all cutoffs and combine
person_level_sleep_all <- bind_rows(
  lapply(cutoffs, function(cut) person_level_sleep.fun(df = sleep_all, cut = cut))
)

# Check result
head(person_level_sleep_all)
tail(person_level_sleep_all)

summary(sleep_all)

```


```{r}
## Sanity check: the sum of mean sleep time and mean sedentary time should be less than or equal to 24 hours for each ID and cutoff (since they are based on the same hourly data)
## The sum of the avg sleep hours and avg wake hours should be less than or equal to 24 hours for each ID and cutoff

person_level_slp_sed_check <- person_level_sed_all %>%
  mutate(
    total_avg_hours = avg_sleep_hours + avg_wake_hours
  )

# Check if any total_time exceeds 24 hours
invalid_time_rows <- person_level_slp_sed_check %>%
  filter(total_avg_hours > 24)

print(paste0("Number of rows where total time exceeds 24 hours: ", nrow(invalid_time_rows)))

head(person_level_slp_sed_check)

```


## Combine person-level sleep and sedentary summaries into one dataframe for modeling

```{r combine into one dataframe}
person_level_slp_sed <- person_level_sed_all %>% 
  left_join(person_level_sleep_all, by=c('ID','cutoff') )

summary(person_level_slp_sed)
na_rows <- person_level_slp_sed %>%
  filter(is.na(mean_sleep_time))

# View them
na_rows

# Count how many
nrow(na_rows)

length(cutoffs)

length(unique(person_level_sleep_all$ID))

```





## Perpare demographic variables for modeling
## Demographic variables are time invariant, so we can just take the first value for each ID. We also create some new variables like age group and BMI category for modeling. We will merge these demographic variables with the person-level sleep and sedentary summaries for modeling.

```{r}
sedentary.df$Dataset <- ifelse(sedentary.df$Dataset == "2011-12", "2011-2012", "2013-2014")
```


```{r}

cl = unique(sedentary.df$ID)
sed_clean1.df <- sedentary.df[sedentary.df$ID %in% cl,]

demo.df <- sed_clean1.df %>% 
  mutate(
    Age_group =ifelse(Age <18, '<18', ifelse(Age>=18 & Age<=34, '18-34',
                                             ifelse(Age>=35 & Age<=44, '35-44',
                                              ifelse(Age>=45 & Age<=64, '45-64',
                                              '>=65')) )),
    BMI= Weight/(Height/100)^2,
    BMI_cat = ifelse(BMI< 25, 'under/normal weight', ifelse(BMI>= 25 & BMI< 30,
                                                            'over weight', 'obesity'))
  )

demo_unique <- demo.df %>%
  group_by(ID) %>%
  summarise(
    Marital_status = first(Marital_status),
    Education = first(Education_coarse),
    Race = first(Race),
    Gender = first(Gender),
    Age = first(Age),
    Weight = first(Weight),
    Height = first(Height),
    Age_group= first(Age_group),
    BMI_cat =first(BMI_cat),
    Exam_weight_recal= first(Exam_weight_recal), 
    PSU=first(PSU), Stratum= first(Stratum),
    .groups = "drop"
  )

head(demo_unique)
```


## Merge the demographic variables with the person-level sleep and sedentary summaries for modeling. This will allow us to include demographic covariates in our models if needed.



```{r, Aggregate to participant level per cutoff, echo=FALSE, warning=FALSE}

df_person <- sed_all %>%
  group_by(ID, cutoff) %>%
  summarise(
    mean_sed_hours = mean(total_sed_hours, na.rm = TRUE),
    mean_sleep_hours = mean(sleep_hrs, na.rm = TRUE),
    .groups = "drop"
  )

head(df_person)
```

```{r Merge with demographic variables & NHANES survey variables}

person_level_demo <- df_person %>%
  left_join(
    demo_unique %>% 
      select(ID, Age_group, BMI_cat, Race, Gender,
             Education, Marital_status,
             Exam_weight_recal, PSU, Stratum),
    by = "ID"
  )



person_level_demo <- person_level_demo %>% 
  mutate(mean_wake_hours = 24 - mean_sleep_hours)

model_df <- person_level_demo %>% 
  left_join(
    sedentary.df %>% 
      select(ID,  Dataset) %>% 
      distinct(ID,  Dataset),
    by = "ID"
  )
tail(model_df)
```


```{r Function to get survey-weighted estimates for each cutoff, echo=FALSE, warning=FALSE}
get_svy_estimates <- function(data_cut) {
  des <- svydesign(
    id = ~PSU,
    strata = ~Stratum,
    weights = ~Exam_weight_recal,
    nest = TRUE,
    data = data_cut
  )
  
  sleep_est <- svymean(~mean_sleep_hours, des, na.rm = TRUE)
  sed_est   <- svymean(~mean_sed_hours, des, na.rm = TRUE)
  
  data.frame(
    cutoff = unique(data_cut$cutoff),
    sleep_mean = coef(sleep_est)[1],
    sleep_se = SE(sleep_est)[1],
    sed_mean = coef(sed_est)[1],
    sed_se = SE(sed_est)[1]
  )
}
```





```{r apply the function to get survey-weighted estimates for each cutoff, echo=FALSE, warning=FALSE}
results <- model_df %>%
  group_split(cutoff) %>%
  map_df(get_svy_estimates)
```

```{r}
head(results)
```


## NHANES self-reported sleep hours(dataset: sleep Disorders)

```{r Get self-reported sleep hours from NHANES for comparison, echo=FALSE, warning=FALSE}
library(haven)
library(dplyr)
library(survey)

## Load and combine NHANES sleep questionnaire data
sleep_2011_12 <- read_xpt("data/self-sleep-2011-12.xpt")
sleep_2013_14 <- read_xpt("data/self-sleep-2013-14.xpt")

self_sleep <- bind_rows(sleep_2011_12, sleep_2013_14) %>%
  transmute(
    ID = SEQN,
    self_reported_slp_hrs = SLD010H
  ) %>%
  ## Remove refused = 77 and don't know = 99
  filter(
    !is.na(self_reported_slp_hrs),
    !(self_reported_slp_hrs %in% c(77, 99))
  )

## Add NHANES survey design variables
self_sleep_weighted <- self_sleep %>%
  left_join(
    demo_unique %>% select(ID, Exam_weight_recal, PSU, Stratum),
    by = "ID"
  ) %>%
  filter(!is.na(Exam_weight_recal))

## Survey design
des_self_sleep <- svydesign(
  id = ~PSU,
  strata = ~Stratum,
  weights = ~Exam_weight_recal,
  nest = TRUE,
  data = self_sleep_weighted
)

## Weighted mean and 95% CI
mean_self_sleep <- svymean(
  ~self_reported_slp_hrs,
  design = des_self_sleep,
  na.rm = TRUE
)

mean_self_sleep
confint(mean_self_sleep)
```



```{r, select cutoff with sleep time closset to the self reported sleep, echo=FALSE, warning=FALSE}
target_sleep <- mean_self_sleep # Replace with the actual self-reported sleep hours from NHANES

results <- results %>%
  mutate(diff = abs(sleep_mean - target_sleep))

best_cutoff <- results %>%
  slice_min(diff, n = 1)

best_cutoff
```

Cutoff Selection Procedure

Self-reported sleep duration was obtained from the NHANES sleep questionnaire, and the survey-weighted mean was estimated as 6.95 hours. This value was used as a population-level calibration target.

For a range of candidate sleep cutoffs, accelerometer data were used to classify each hour as sleep or wake. Daily sleep hours were computed for each participant, averaged across days, and then summarized at the population level using survey weights, accounting for stratification and clustering.

The optimal cutoff was defined as the value that minimized the absolute difference between the survey-weighted mean accelerometer-derived sleep duration and the self-reported mean. Based on this criterion, a cutoff of 91% was selected, yielding an estimated mean sleep duration of 6.91 hours (SE = 0.053), which is closest to the self-reported benchmark.

Using this cutoff, the corresponding survey-weighted estimate of sedentary behavior was 11.35 hours/day (SE = 0.044).



## Data for Estimating sedentary time in wake hours and sleep time for different cutoffs


```{r, check the data for modeling, echo=FALSE, warning=FALSE}
head(person_level_demo)
summary(person_level_demo)
length(unique(person_level_demo$ID))
```


## Example model to examine the relationship between cutoff and sleep/sedentary time while accounting for repeated measures within individuals (ID) using Generalized Additive Mixed Models (GAMMs). We will use the `gamm4` package for this purpose. The models will include a random intercept for ID to account for within-person correlation across different cutoffs.

## This is optional analysis to explore the relationship between cutoff and sleep/sedentary time, and to visualize the predicted curves. The main analysis will be based on survey-weighted estimates, but this can provide additional insights into the shape of the relationship across cutoffs.

```{r SLEEP & SEDENTARY GAMM FIT }
library(emmeans)
library(gamm4)

weights <- c(person_level_demo %>% 
  select(Exam_weight_recal))
## scale weights to be between 0 and 1
weig <- (weights$Exam_weight_recal)/ sum((weights$Exam_weight_recal))

# Fit sleep model
sleep.fit <- bam(
  avg_sleep_hours ~ cutoff + s(ID, bs="re"),
  data    = person_level_demo,
  weights = weig,
  discrete = TRUE
)


summary(sleep.fit)
plot(sleep.fit, pages=1, shade=TRUE)


## Fit sedentary time model
sed.fit <- bam(
  mean_sed_time ~ cutoff + s(ID, bs="re"),
  data    = person_level_demo,
  weights = weig,
  discrete = TRUE
)

summary(sed.fit)


## Make predictions and plot
# Pick a valid ID
one_id <- person_level_demo$ID[6]

# Create the cutoff grid with a valid ID
newdat <- data.frame(
  cutoff = seq(min(person_level_demo$cutoff),
               max(person_level_demo$cutoff),
              2),
  ID = one_id
)

# Predict population-level smooth
# Sleep
pred_slp <- predict(
  sleep.fit,
  newdata = newdat,
  se.fit = TRUE,
  exclude = "s(ID)"
)

## Sedentary
pred_sed <- predict(
  sed.fit,
  newdata = newdat,
  se.fit = TRUE,
  exclude = "s(ID)"
)

slp_df <- newdat
slp_df$fit <- pred_slp$fit
slp_df$se  <- pred_slp$se.fit
slp_df$type <- rep("sleep", length(slp_df$fit))

sd_df <- newdat
sd_df$fit <- pred_sed$fit
sd_df$se  <- pred_sed$se.fit
sd_df$type <- rep("sed", length(sd_df$fit))

final_df <- rbind(slp_df, sd_df)

pg<- ggplot(final_df, aes(x = cutoff, y = fit, color = type, fill = type)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = fit - 1.96 * se,
                  ymax = fit + 1.96 * se),
              alpha = 0.5, color = NA) +
  labs(
    x = "% Sleep Cutoff",
    y = "Predicted Hours per Day (Weighted)",
    title = NULL,
    color = NULL,
    fill  = NULL
  ) +
  # --- X-axis scale controls ---
  scale_x_continuous(
    name   = "% Sleep Cutoff",
    limits = c(5, 95),        # adjust as needed
    breaks = seq(5, 95, 5)  # adjust tick spacing
  ) +

  # --- Y-axis scale controls ---
  scale_y_continuous(
    name   = "hour/Day (Weighted)",
    limits = c(0, 20),        # adjust as needed
    breaks = seq(0, 20, 1)    # adjust tick spacing
  ) +
  scale_color_manual(values = c("sleep" = "purple", "sed" = "black")) +
  scale_fill_manual(values  = c("sleep" = "skyblue", "sed" = "skyblue")) +
  theme_classic(base_size = 14) +
  theme(legend.position = "bottom")

pg
```












## Using the percentage sleep cut-off of 91% to get sedentary estimates.
##  This will allow us to get the survey-weighted estimates of sedentary time in wake hours and sleep time for this specific cutoff that matches the self-reported sleep hours.

```{r}
model_df_91 <- model_df %>% filter(cutoff == best_cutoff$cutoff)
head(model_df_91)
```

```{r}
length(unique(model_df_91$ID))
```


```{r}
library(survey)

des91 <- svydesign(
  id = ~PSU,
  strata = ~Stratum,
  weights = ~Exam_weight_recal,
  nest = TRUE,
  data = model_df_91
)

wake_mean <- svymean(~mean_wake_hours, des91, na.rm = TRUE)

wake_mean
confint(wake_mean)
```


## Distribution of sedentary time

```{r}
ggplot(model_df_91, aes(x = mean_sed_hours)) +
  geom_histogram(
    breaks = c(0, 1, 2, 3, 4, 5,6,7,8,9,10,11,12,13,14,15, Inf),
    closed = "right"
  ) +
  labs(
    x = "Daily sedentary time (hours/day)",
    y = "Number of participants"
  ) +
  theme_classic(base_size = 14)
```


```{r}
library(dplyr)
library(ggplot2)
library(scales)

## Create sedentary-time bins
df_plot <- model_df_91 %>%
  distinct(ID, .keep_all = TRUE) %>%
  mutate(
    sed_bin = cut(
      mean_sed_hours,
      breaks = c(0, 4, seq(6, 16, by = 2), Inf),
      labels = c(
        "0–4", "4–6", "6–8", "8–10",
        "10–12", "12–14", "14–16", "16+"
      ),
      include.lowest = TRUE,
      right = FALSE
    )
  ) %>%
  filter(!is.na(sed_bin))

## Count participants and proportions
plot_df <- df_plot %>%
  group_by(sed_bin) %>%
  summarise(
    n = n_distinct(ID),
    prop = n / n_distinct(df_plot$ID),
    .groups = "drop"
  )

## Plot for small panel
sed_dist_plot <- ggplot(plot_df, aes(x = sed_bin, y = prop)) +
  geom_col(width = 0.65, color = "black", linewidth = 0.2) +
  geom_text(
    aes(label = n),
    vjust = -0.3,
    size = 2.2,
    fontface = "bold"
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.18))
  ) +
  labs(
    x = "Sedentary time (h/day)",
    y = "Participants (%)"
  ) +
  theme_classic(base_size = 7) +
  theme(
    axis.title.x = element_text(
      face = "bold",
      size = 7,
      margin = margin(t = 1)
    ),
    axis.title.y = element_text(
      face = "bold",
      size = 7,
      margin = margin(r = 1)
    ),
    axis.text.x = element_text(
      face = "bold",
      size = 5.5,
      angle = 45,
      hjust = 1,
      vjust = 1
    ),
    axis.text.y = element_text(
      face = "bold",
      size = 5.5
    ),
    plot.margin = margin(2, 2, 2, 2)
  )

## Display plot
sed_dist_plot

## Save exactly at panel size
ggsave(
  filename = "figures/sedentary_distribution_small_panel.png",
  plot = sed_dist_plot,
  width = 2.67,
  height = 1.8,
  units = "in",
  dpi = 600
)
```


```{r}
head(model_df_91)
```


```{r}
library(dplyr)
library(ggplot2)
library(scales)
library(tidyr)
library(patchwork)

## Create 2-hour sedentary-time bins
df_plot <- model_df_91 %>%
  distinct(ID, .keep_all = TRUE) %>%
  mutate(
    sed_bin = cut(
      mean_sed_hours,
      breaks = c(0,4,seq(6, 16, by = 2), Inf),
      labels = c(
        "0–4", "4–6", "6–8", "8–10",
        "10–12", "12–14", "14–16", "16+"
      ),
      include.lowest = TRUE,
      right = FALSE
    )
  ) %>%
  filter(!is.na(sed_bin))

## Count participants and proportions for left panel
plot_df <- df_plot %>%
  group_by(sed_bin) %>%
  summarise(
    n = n_distinct(ID),
    prop = n / n_distinct(df_plot$ID),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("n=", scales::comma(n), "\n", scales::percent(prop, accuracy = 0.1))
  )

## Left panel: distribution of participants by SB group
left_plot <- ggplot(plot_df, aes(x = sed_bin, y = prop)) +
  geom_col(width = 0.72, fill = "#4E79A7", color = "black", linewidth = 0.35) +
  geom_text(
    aes(label = label),
    vjust = -0.25,
    size = 4.2,
    fontface = "bold",
    lineheight = 0.9
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.18))
  ) +
  labs(
    x = "Daily sedentary time (hours/day)",
    y = "Participants (%)"
  ) +
  coord_cartesian(clip = "off") +
  theme_bw(base_size = 17) +
  theme(
    axis.title = element_text(face = "bold", size = 16),
    axis.text = element_text(color = "black", size = 13),
    axis.text.x = element_text(angle = 35, hjust = 1, vjust = 1),
    axis.ticks = element_line(color = "black", linewidth = 0.45),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey88", linewidth = 0.35),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", linewidth = 0.7),
    plot.margin = margin(10, 16, 10, 10)
  ) +
  labs(tag = "A")

## Right panel: 24-hour composition in hours,
## with percentages calculated for labels
right_df <- df_plot %>%
  group_by(sed_bin) %>%
  summarise(
    SB = mean(mean_sed_hours, na.rm = TRUE),
    Sleep = mean(mean_sleep_hours, na.rm = TRUE),
    Activity = mean(24 - mean_sed_hours - mean_sleep_hours, na.rm = TRUE),
    .groups = "drop"
  )

## Convert to long format
right_df_long <- right_df %>%
  pivot_longer(
    cols = c(SB, Sleep, Activity),
    names_to = "behavior",
    values_to = "hours"
  ) %>%
  mutate(
    behavior = factor(
      behavior,
      levels = c("Sleep", "SB", "Activity"),
      labels = c("Sleep/nonwear", "Sedentary", "Other waking time")
    ),
    percent_day = hours / 24,
    percent_label = ifelse(hours >= 1.2, percent(percent_day, accuracy = 1), ""),
    label_color = ifelse(behavior == "Sedentary", "white", "black")
  )

right_label_df <- right_df %>%
  transmute(
    sed_bin,
    `Other waking time` = Activity,
    Sedentary = SB,
    `Sleep/nonwear` = Sleep
  ) %>%
  pivot_longer(
    cols = c(`Other waking time`, Sedentary, `Sleep/nonwear`),
    names_to = "behavior",
    values_to = "hours"
  ) %>%
  group_by(sed_bin) %>%
  mutate(
    x_mid = cumsum(hours) - hours / 2,
    percent_label = ifelse(hours >= 1.2, percent(hours / 24, accuracy = 1), ""),
    label_color = ifelse(behavior %in% c("Other waking time", "Sedentary"), "white", "black")
  ) %>%
  ungroup()

## Right panel: stacked horizontal bar plot
right_plot <- ggplot(
  right_df_long,
  aes(x = hours, y = sed_bin, fill = behavior)
) +
  geom_col(width = 0.72, color = "white", linewidth = 0.45) +
  geom_text(
    data = right_label_df,
    aes(x = x_mid, y = sed_bin, label = percent_label, color = label_color),
    inherit.aes = FALSE,
    size = 4.2,
    fontface = "bold"
  ) +
  scale_fill_manual(
    values = c(
      "Sleep/nonwear" = "#9ECAE1",
      "Sedentary" = "#4E79A7",
      "Other waking time" = "#59A14F"
    )
  ) +
  scale_color_identity(guide = "none") +
  guides(fill = guide_legend(reverse = TRUE)) +
  scale_x_continuous(
    breaks = seq(0, 24, by = 4),
    expand = expansion(mult = c(0, 0.02))
  ) +
  coord_cartesian(xlim = c(0, 24), clip = "off") +
  labs(
    x = "Hours per day",
    y = "Daily sedentary time (hours/day)",
    fill = NULL
  ) +
  theme_bw(base_size = 17) +
  theme(
    axis.title = element_text(face = "bold", size = 16),
    axis.text = element_text(color = "black", size = 13),
    axis.ticks = element_line(color = "black", linewidth = 0.45),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "grey88", linewidth = 0.35),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", linewidth = 0.7),
    legend.position = "bottom",
    legend.text = element_text(size = 13),
    legend.key.width = unit(1.1, "cm"),
    plot.margin = margin(10, 10, 10, 16)
  ) +
  labs(tag = "B")

## Combine both panels
combined_plot <- left_plot + right_plot +
  plot_layout(widths = c(1, 1.25), guides = "collect") &
  theme(
    legend.position = "bottom",
    plot.tag = element_text(face = "bold", size = 19),
    plot.tag.position = c(0.02, 0.98)
  )

## Display plot
combined_plot

## Save plot
ggsave(
  "figures/sedentary_distribution_with_24h_composition.png",
  combined_plot,
  width = 12.5,
  height = 6.2,
  units = "in",
  dpi = 600,
  bg = "white"
)

ggsave(
  "figures/sedentary_distribution_with_24h_composition.pdf",
  combined_plot,
  width = 12.5,
  height = 6.2,
  units = "in",
  bg = "white"
)
```




```{r}
plot_df %>%
  summarise(total_n = sum(n))
df_plot %>%
  count(sed_bin)

df_plot %>%
  summarise(
    total_rows = n(),
    total_ids = n_distinct(ID)
  )
```


```{r}
library(dplyr)
library(survey)
library(ggplot2)
library(scales)

## 1. Create sedentary-time categories
df_plot <- model_df_91 %>%
  mutate(
    sed_cat = case_when(
      mean_sed_hours <= 4  ~ "0-4",
      mean_sed_hours <= 7  ~ "5-7",
      mean_sed_hours <= 10 ~ "8-10",
      mean_sed_hours <= 12  ~ "11-12",
       mean_sed_hours > 12 ~ "13+",
      TRUE ~ NA_character_
    ),
    sed_cat = factor(sed_cat, levels = c("0-4", "5-7", "8-10", "11-12", "13+"))
  ) %>%
  filter(!is.na(sed_cat))

## 2. Unweighted counts
count_df <- df_plot %>%
  group_by(sed_cat) %>%
  summarise(n = n_distinct(ID), .groups = "drop")

## 3. Survey-weighted proportions
des91_cat <- svydesign(
  id = ~PSU,
  strata = ~Stratum,
  weights = ~Exam_weight_recal,
  nest = TRUE,
  data = df_plot
)

prop_est <- svymean(~sed_cat, des91_cat, na.rm = TRUE)

prop_df <- data.frame(
  sed_cat = names(coef(prop_est)),
  proportion = as.numeric(coef(prop_est))
) %>%
  mutate(
    sed_cat = gsub("^sed_cat", "", sed_cat),
    sed_cat = factor(sed_cat, levels = c("0-4", "5-7", "8-10", "11-12", "13+"))
  )

## 4. Combine counts and proportions
plot_df <- prop_df %>%
  left_join(count_df, by = "sed_cat") %>%
  arrange(sed_cat)

## 5. Plot
ggplot(plot_df, aes(x = sed_cat, y = proportion)) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = n),
    vjust = -0.4,
    size = 5
  ) +
  scale_x_discrete(limits = c("0-4", "5-7", "8-10", "11-12","13+")) +
  scale_y_continuous(
    labels = percent_format(),
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    x = "Daily sedentary time (hours)",
    y = "Proportion of population"
  ) +
  theme_classic(base_size = 14)


```




```{r}
library(ggplot2)
library(scales)

ggplot(plot_df, aes(x = sed_cat, y = proportion)) +
  geom_col(width = 0.6) +
  
  ## counts on top
  geom_text(
    aes(label = n),
    vjust = -0.4,
    size = 5
  ) +
  
  scale_y_continuous(
    labels = percent_format(),
    expand = expansion(mult = c(0, 0.1))
  ) +
  
  labs(
    x = "Daily sedentary time (hours)",
    y = "Proportion of population"
  ) +
  
  theme_classic(base_size = 14)

```




## Compare mean sed estimate across NHANES circles for the chosen cutoff


```{r}
library(survey)

des <- svydesign(
  id = ~PSU,
  strata = ~Stratum,
  weights = ~Exam_weight_recal,
  nest = TRUE,
  data = model_df_91
)
svyby(
  ~mean_sed_hours,
  ~Dataset,
  des,
  svymean,
  na.rm = TRUE
)
```

```{r}
svyglm(
  mean_sed_hours ~ Dataset,
  design = des
)
```

## Overall mean sedentary time for the chosen cutoff

```{r}
overall <- svymean(~mean_sed_hours, des, na.rm = TRUE)
overall
confint(overall)
```

## Demographic subgroup comparisons for the chosen cutoff
### Age group comparisons

```{r}
svyby(
  ~mean_sed_hours,
  ~Age_group,
  des,
  svymean,
  na.rm = TRUE
)
```

## Gender comparisons
```{r}
svyby(
  ~mean_sed_hours,
  ~Gender,
  des,
  svymean,
  na.rm = TRUE
)
```

## BMI category comparisons

```{r}
svyby(
  ~mean_sed_hours,
  ~BMI_cat,
  des,
  svymean,
  na.rm = TRUE
)
```


### Race comparisons

```{r}
svyby(
  ~mean_sed_hours,
  ~Race,
  des,
  svymean,
  na.rm = TRUE
)
```


```{r}
model <- svyglm(
  mean_sed_hours ~ Age_group + Gender + BMI_cat + Race,
  design = des
)

summary(model)
```

Using the calibrated 91% sleep cutoff, the survey-weighted mean sedentary time was 11.35 hours/day (95% CI: 11.26, 11.44). Estimates were similar across the two NHANES cycles, with means of 11.46 hours/day in 2011–2012 and 11.25 hours/day in 2013–2014, supporting the pooling of cycles. Sedentary time increased with age, from 10.38 hours/day among participants under 18 years to 13.16 hours/day among those aged 65 years and older. Sedentary time also increased across BMI categories, from 10.79 hours/day among under/normal weight participants to 12.08 hours/day among participants with obesity. Males had slightly higher sedentary time than females. These patterns were largely consistent in adjusted survey-weighted regression models.


## Create a table summarizing the mean sedentary time (with 95% CI) and unweighted N (%) for each age group and gender, using the survey design to get the estimates for sedentary time and the original dataset to get the unweighted N (%). We will create two tables: one for the mean sedentary time and one for the N (%), and then we can combine them into one table if needed.

```{r}

library(dplyr)
library(survey)
#library(knitr)

## Survey design
des91 <- svydesign(
  id = ~PSU,
  strata = ~Stratum,
  weights = ~Exam_weight_recal,
  nest = TRUE,
  data = model_df_91
)

## Function: mean and 95% CI
get_est <- function(design_obj) {
  m <- svymean(~mean_sed_hours, design_obj, na.rm = TRUE)
  ci <- confint(m)
  
  sprintf("%.3f (%.3f, %.3f)", coef(m)[1], ci[1, 1], ci[1, 2])
}

## Function: unweighted N (%)
get_n <- function(df) {
  n <- n_distinct(df$ID)
  pct <- 100 * n / n_distinct(model_df_91$ID)
  sprintf("%d (%.1f%%)", n, pct)
}

## Age order
age_levels <- c("<18", "18-34", "35-44", "45-64", ">=65")

## Mean sedentary time table: age rows, gender columns
table_age_gender <- lapply(age_levels, function(a) {
  
  data.frame(
    Age_group = a,
    All = get_est(subset(des91, Age_group == a)),
    Male = get_est(subset(des91, Age_group == a & Gender == "Male")),
    Female = get_est(subset(des91, Age_group == a & Gender == "Female"))
  )
}) %>%
  bind_rows()

## Add overall row
overall_row <- data.frame(
  Age_group = "All",
  All = get_est(des91),
  Male = get_est(subset(des91, Gender == "Male")),
  Female = get_est(subset(des91, Gender == "Female"))
)

table_age_gender <- bind_rows(overall_row, table_age_gender)

## N (%) table
n_table <- lapply(age_levels, function(a) {
  
  df_sub <- model_df_91 %>% filter(Age_group == a)
  
  data.frame(
    Age_group = a,
    All = get_n(df_sub),
    Male = get_n(df_sub %>% filter(Gender == "Male")),
    Female = get_n(df_sub %>% filter(Gender == "Female"))
  )
}) %>%
  bind_rows()

## Add overall N row
n_overall_row <- data.frame(
  Age_group = "All",
  All = get_n(model_df_91),
  Male = get_n(model_df_91 %>% filter(Gender == "Male")),
  Female = get_n(model_df_91 %>% filter(Gender == "Female"))
)

n_table <- bind_rows(n_overall_row, n_table)

## Print mean table
knitr::kable(
  table_age_gender,
  col.names = c(
    "Age group (years)",
    "All: Mean (95% CI)",
    "Male: Mean (95% CI)",
    "Female: Mean (95% CI)"
  ),
  caption = "Weighted mean sedentary time (hours/day), by age group and gender, United States, NHANES 2011–2014."
)

## Print N table
knitr::kable(
  n_table,
  col.names = c(
    "Age group (years)",
    "All: N (%)",
    "Male: N (%)",
    "Female: N (%)"
  ),
  caption = "Analytic sample size by age group and gender, NHANES 2011–2014."
)
```


## Create a table summarizing the mean sedentary time (with 95% CI) and unweighted N (%) for each BMI category by race.

```{r}
library(dplyr)
library(survey)
#library(knitr)

## Survey design
des91 <- svydesign(
  id = ~PSU,
  strata = ~Stratum,
  weights = ~Exam_weight_recal,
  nest = TRUE,
  data = model_df_91
)

## Function: mean + CI
get_est <- function(design_obj) {
  m <- svymean(~mean_sed_hours, design_obj, na.rm = TRUE)
  ci <- confint(m)
  
  sprintf("%.2f (%.2f, %.2f)", coef(m)[1], ci[1,1], ci[1,2])
}

## Optional: N (%)
get_n <- function(df) {
  n <- n_distinct(df$ID)
  pct <- 100 * n / n_distinct(model_df_91$ID)
  sprintf("%d (%.1f%%)", n, pct)
}

## Order
race_levels <- c(
  "Mexican American",
  "NH Asian",
  "NH Black",
  "NH White",
  "Other Hispanic",
  "Other/Multi"
)

bmi_levels <- c("under/normal weight", "over weight", "obesity")

## Build mean table
table_race_bmi <- lapply(race_levels, function(race) {
  
  data.frame(
    Race = race,
    `Under/Normal weight` = get_est(subset(des91, Race == race & BMI_cat == "under/normal weight")),
    `Overweight`          = get_est(subset(des91, Race == race & BMI_cat == "over weight")),
    `Obesity`             = get_est(subset(des91, Race == race & BMI_cat == "obesity"))
  )
}) %>% bind_rows()

## Add overall row
overall_row <- data.frame(
  Race = "All",
  `Under/Normal weight` = get_est(subset(des91, BMI_cat == "under/normal weight")),
  `Overweight`          = get_est(subset(des91, BMI_cat == "over weight")),
  `Obesity`             = get_est(subset(des91, BMI_cat == "obesity"))
)

table_race_bmi <- bind_rows(overall_row, table_race_bmi)

## Display
knitr::kable(
  table_race_bmi,
  caption = "Weighted mean sedentary time (hours/day) by race/ethnicity and BMI category, NHANES 2011–2014."
)
```


```{r}
n_table_race_bmi <- lapply(race_levels, function(race) {
  
  df_sub <- model_df_91 %>% filter(Race == race)
  
  data.frame(
    Race = race,
    `Under/Normal weight` = get_n(df_sub %>% filter(BMI_cat == "under/normal weight")),
    `Overweight`          = get_n(df_sub %>% filter(BMI_cat == "over weight")),
    `Obesity`             = get_n(df_sub %>% filter(BMI_cat == "obesity"))
  )
}) %>% bind_rows()

n_overall <- data.frame(
  Race = "All",
  `Under/Normal weight` = get_n(filter(model_df_91, BMI_cat == "under/normal weight")),
  `Overweight`          = get_n(filter(model_df_91, BMI_cat == "over weight")),
  `Obesity`             = get_n(filter(model_df_91, BMI_cat == "obesity"))
)

n_table_race_bmi <- bind_rows(n_overall, n_table_race_bmi)

knitr::kable(
  n_table_race_bmi,
  caption = "Sample size by race/ethnicity and BMI category, NHANES 2011–2014."
)
```





```{r, cutoff dataset}
sleep_aware_df <- function(df, cutoffs) {

  all_IDs <- unique(df$ID)
  results_sed <- map(cutoffs, function(cut) {
    ## Show progress
    #print(paste0("Processing cutoff: ", cut, "% sleep"))
    # ----- 1) Sedentary (wake) summary -----
    sed_df <- df %>%
      filter(percent_sleep < cut) 
list(sed_df = sed_df)
  })
  
  # Combine results across all cutoffs
    sedentary_big <- bind_rows(map(results_sed, "sed_df"))
    
  return(sed_summ_df = sedentary_big)
}

cutoffs <- best_cutoff$cutoff
slp_aware_sed.df <- sleep_aware_df(sleep_sed.df, cutoffs = cutoffs)

```

```{r, view sleep aware sed df}
head(slp_aware_sed.df)
dim(slp_aware_sed.df)
```

## Plot hourly estimates of sedentary time across the day for the chosen cutoff, stratified by weekday vs weekend, using survey-weighted estimates. compare to when sleep is not taken into account (i.e. using the original hourly dataset without applying the sleep cutoff). This will allow us to see how the pattern of sedentary time across the day changes when we account for sleep using the calibrated cutoff.


```{r}
make_hourly_plot <- function(df, yvar, ylab, title_label, y_limits) {
  
  df <- df %>%
    mutate(DayType = factor(DayType, levels = c("Weekend", "Weekday")))
  
  des <- svydesign(
    id = ~PSU,
    strata = ~Stratum,
    weights = ~Exam_weight_recal,
    nest = TRUE,
    data = df
  )
  
  hourly_est <- svyby(
    as.formula(paste0("~", yvar)),
    by = ~DayType + Hour,
    design = des,
    FUN = svymean,
    vartype = "ci",
    na.rm = TRUE
  ) %>%
    rename(estimate = !!sym(yvar))
  
  fig1_colors <- c("Weekend" = "#0072B2", "Weekday" = "#D55E00")
  
  ggplot(hourly_est, aes(x = Hour, y = estimate, group = DayType)) +
    geom_ribbon(
      aes(ymin = ci_l, ymax = ci_u, fill = DayType),
      alpha = 0.28,
      color = NA
    ) +
    geom_line(aes(y = ci_l, color = DayType), linewidth = 0.35, alpha = 0.65) +
    geom_line(aes(y = ci_u, color = DayType), linewidth = 0.35, alpha = 0.65) +
    geom_line(aes(color = DayType, linetype = DayType), linewidth = 1.35) +
    geom_point(
      aes(color = DayType, shape = DayType),
      size = 3.3,
      stroke = 1.25,
      fill = "white"
    ) +
    scale_color_manual(values = fig1_colors) +
    scale_fill_manual(values = fig1_colors) +
    scale_linetype_manual(
      values = c("Weekend" = "solid", "Weekday" = "dashed")
    ) +
    scale_shape_manual(
      values = c("Weekend" = 16, "Weekday" = 1)
    ) +
    scale_x_continuous(
      breaks = seq(0, 24, 4),
      minor_breaks = seq(1, 24, 1),
      limits = c(1, 24),
      expand = expansion(mult = c(0.01, 0.02))
    ) +
    scale_y_continuous(
      limits = y_limits,
      labels = scales::percent_format(scale = 1),
      breaks = seq(y_limits[1], y_limits[2], 10),
      expand = expansion(mult = c(0.02, 0.04))
    ) +
    labs(
      x = "Hour of day",
      y = ylab,
      title = title_label,
      color = NULL,
      fill = NULL,
      linetype = NULL,
      shape = NULL
    ) +
    guides(fill = "none") +
    theme_bw(base_size = 15) +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5, margin = margin(b = 8)),
      axis.title = element_text(face = "bold", size = 14),
      axis.text = element_text(color = "black", size = 12),
      axis.ticks = element_line(color = "black", linewidth = 0.45),
      axis.ticks.length = unit(3, "pt"),
      panel.grid.major = element_line(color = "grey88", linewidth = 0.35),
      panel.grid.minor.x = element_line(color = "grey94", linewidth = 0.25),
      panel.grid.minor.y = element_blank(),
      panel.border = element_rect(color = "black", linewidth = 0.7),
      legend.position = "bottom",
      legend.text = element_text(size = 13),
      legend.key.width = unit(1.15, "cm"),
      legend.margin = margin(t = 2, b = 2),
      plot.margin = margin(8, 10, 8, 10)
    )
}
```



```{r}
hourly_plot <- function(df, yvar, ylab, title_label = NULL, y_limits) {
  
  des <- svydesign(
    id = ~PSU,
    strata = ~Stratum,
    weights = ~Exam_weight_recal,
    nest = TRUE,
    data = df
  )
  
  hourly_est <- svyby(
    as.formula(paste0("~", yvar)),
    by = ~Hour,
    design = des,
    FUN = svymean,
    vartype = "ci",
    na.rm = TRUE
  ) %>%
    rename(estimate = !!sym(yvar))
  
  ggplot(hourly_est, aes(x = Hour, y = estimate)) +
    geom_ribbon(
      aes(ymin = ci_l, ymax = ci_u),
      fill = "grey80",
      alpha = 0.6,
      color = NA
    ) +
    geom_line(linewidth = 0.7, color = "black") +
    geom_point(size = 0.8, color = "black") +
    scale_x_continuous(
      breaks = seq(0, 24, 6),
      limits = c(1, 24)
    ) +
    scale_y_continuous(
      limits = y_limits,
      labels = scales::percent_format(scale = 1)
    ) +
    labs(
      x = "Hour of day",
      y = ylab,
      title = title_label
    ) +
    theme_classic(base_size = 8) +
    theme(
      plot.title = element_text(
        face = "bold",
        size = 8,
        hjust = 0.5,
        margin = margin(b = 2)
      ),
      axis.title.x = element_text(
        face = "bold",
        size = 7,
        margin = margin(t = 1)
      ),
      axis.title.y = element_text(
        face = "bold",
        size = 7,
        margin = margin(r = 1)
      ),
      axis.text.x = element_text(
        face = "bold",
        size = 6
      ),
      axis.text.y = element_text(
        face = "bold",
        size = 6
      ),
      plot.margin = margin(2, 2, 2, 2)
    )
}
```

```{r}
p_hourly <- hourly_plot(
  df = slp_aware_sed.df,
  yvar = "percent_sedentary_wrist",
  ylab = "Sedentary (%)",
  title_label = "Waking sedentary (sleep >= 91% removed)",
  y_limits = c(40, 100)
)

p_hourly

ggsave(
  filename = "figures/hourly_plot_small_panel.png",
  plot = p_hourly,
  width = 2.67,
  height = 1.8,
  units = "in",
  dpi = 600
)
```


```{r}
p_left <- make_hourly_plot(
  df = sleep_sed.df,
  yvar = "percent_sedentary_wrist",
  ylab = "Hourly sitting (%)",
  title_label = "CHAP sitting, sleep included",
  y_limits = c(50, 100)
) +
  labs(tag = "A")

p_right <- make_hourly_plot(
  df = slp_aware_sed.df,
  yvar = "percent_sedentary_wrist",
  ylab = "Hourly waking sedentary (%)",
  title_label = "Waking sedentary, sleep removed",
  y_limits = c(50, 100)
) +
  labs(tag = "B")

fig1 <- p_left + p_right +
  patchwork::plot_layout(guides = "collect", widths = c(1, 1)) &
  theme(
    legend.position = "bottom",
    plot.tag = element_text(face = "bold", size = 19),
    plot.tag.position = c(0.02, 0.98)
  )

fig1
```
Figure 1. Hourly percentage of sitting predicted by CHAP before sleep removal (left) and hourly percentage of waking sedentary time after excluding hours with sleep percentage greater than 91% (right). Shaded bands show 95% confidence intervals. Solid lines with filled points represent weekend estimates, and dashed lines with open points represent weekday estimates.

```{r}
ggsave(
  "figures/fig1.png",
  plot = fig1,
  width = 10,
  height = 5.8,
  units = "in",
  dpi = 600,
  bg = "white"
)

ggsave(
  "figures/fig1.pdf",
  plot = fig1,
  width = 10,
  height = 5.8,
  units = "in",
  bg = "white"
)
```


## Figure 5. Sleep-cutoff calibration and sedentary-time sensitivity. Survey-weighted mean accelerometer-derived sleep duration and waking sedentary time are shown across candidate sleep-percentage cutoffs. Shaded bands indicate 95% confidence intervals. The vertical dashed line marks the selected 91% cutoff calibrated to the self-reported sleep-duration benchmark.


```{r figure-5-cutoff-calibration-sensitivity, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(survey)
library(patchwork)

selected_cutoff <- as.numeric(best_cutoff$cutoff[1])
self_sleep_mean <- as.numeric(coef(mean_self_sleep)[1])
self_sleep_ci <- as.numeric(confint(mean_self_sleep)[1, ])

figure5_df <- results %>%
  mutate(
    sleep_ci_l = sleep_mean - qnorm(0.975) * sleep_se,
    sleep_ci_u = sleep_mean + qnorm(0.975) * sleep_se,
    sed_ci_l = sed_mean - qnorm(0.975) * sed_se,
    sed_ci_u = sed_mean + qnorm(0.975) * sed_se
  )

selected_row <- figure5_df %>%
  filter(cutoff == selected_cutoff) %>%
  slice(1)

fig5_theme <- theme_classic(base_size = 18) +
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0),
    plot.tag = element_text(face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),
    axis.text = element_text(face = "bold", size = 13, color = "black"),
    axis.line = element_line(linewidth = 1.1, color = "black"),
    axis.ticks = element_line(linewidth = 1.1, color = "black"),
    axis.ticks.length = unit(0.16, "cm"),
    legend.position = "none",
    plot.margin = margin(8, 12, 8, 16)
  )

sleep_panel <- ggplot(figure5_df, aes(x = cutoff, y = sleep_mean)) +
  geom_ribbon(
    aes(ymin = sleep_ci_l, ymax = sleep_ci_u),
    fill = "#D9E8F5",
    alpha = 0.95
  ) +
  geom_line(color = "#2F5F98", linewidth = 1.55) +
  geom_point(
    data = selected_row,
    aes(y = sleep_mean),
    color = "#2F5F98",
    fill = "white",
    shape = 21,
    stroke = 1.3,
    size = 3.8
  ) +
  geom_hline(
    yintercept = self_sleep_mean,
    color = "#8B1E2D",
    linewidth = 1.1,
    linetype = "dashed"
  ) +
  geom_vline(
    xintercept = selected_cutoff,
    color = "#8B1E2D",
    linewidth = 1.1,
    linetype = "dashed"
  ) +
  annotate(
    "rect",
    xmin = selected_cutoff - 1.25,
    xmax = selected_cutoff + 1.25,
    ymin = -Inf,
    ymax = Inf,
    fill = "#8B1E2D",
    alpha = 0.08
  ) +
  annotate(
    "text",
    x = selected_cutoff - 1.4,
    y = self_sleep_mean,
    label = "Self-reported sleep mean",
    hjust = 1,
    vjust = -0.5,
    color = "#8B1E2D",
    fontface = "bold",
    size = 4.5
  ) +
  annotate(
    "text",
    x = selected_cutoff,
    y = selected_row$sleep_mean,
    label = paste0(selected_cutoff, "%"),
    hjust = -0.2,
    vjust = 1.7,
    color = "#8B1E2D",
    fontface = "bold",
    size = 4.6
  ) +
  scale_x_continuous(breaks = seq(30, 100, 10), limits = c(25, 100)) +
  scale_y_continuous(
    limits = range(c(figure5_df$sleep_ci_l, figure5_df$sleep_ci_u, self_sleep_ci), na.rm = TRUE) +
      c(-0.15, 0.25),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    tag = "A",
    title = "Sleep-duration calibration",
    x = "Sleep percentage cutoff",
    y = "Mean sleep duration (h/day)"
  ) +
  fig5_theme

sed_panel <- ggplot(figure5_df, aes(x = cutoff, y = sed_mean)) +
  geom_ribbon(
    aes(ymin = sed_ci_l, ymax = sed_ci_u),
    fill = "#E8DFF4",
    alpha = 0.95
  ) +
  geom_line(color = "#7B3FC6", linewidth = 1.55) +
  geom_point(
    data = selected_row,
    aes(y = sed_mean),
    color = "#7B3FC6",
    fill = "white",
    shape = 21,
    stroke = 1.3,
    size = 3.8
  ) +
  geom_vline(
    xintercept = selected_cutoff,
    color = "#8B1E2D",
    linewidth = 1.1,
    linetype = "dashed"
  ) +
  annotate(
    "rect",
    xmin = selected_cutoff - 1.25,
    xmax = selected_cutoff + 1.25,
    ymin = -Inf,
    ymax = Inf,
    fill = "#8B1E2D",
    alpha = 0.08
  ) +
  annotate(
    "text",
    x = selected_cutoff,
    y = selected_row$sed_mean,
    label = paste0(round(selected_row$sed_mean, 2), " h/day"),
    hjust = 1.08,
    vjust = -1.0,
    color = "#7B3FC6",
    fontface = "bold",
    size = 4.6
  ) +
  annotate(
    "text",
    x = selected_cutoff,
    y = max(figure5_df$sed_ci_u, na.rm = TRUE),
    label = paste0(selected_cutoff, "% cutoff"),
    hjust = 1.05,
    vjust = 1.3,
    color = "#8B1E2D",
    fontface = "bold",
    size = 4.6
  ) +
  scale_x_continuous(breaks = seq(30, 100, 10), limits = c(25, 100)) +
  scale_y_continuous(
    limits = range(c(figure5_df$sed_ci_l, figure5_df$sed_ci_u), na.rm = TRUE) + c(-0.12, 0.18),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    tag = "B",
    title = "Sedentary-time sensitivity",
    x = "Sleep percentage cutoff",
    y = "Mean sedentary time (h/day)"
  ) +
  fig5_theme

cutplt <- sleep_panel / sed_panel +
  plot_layout(heights = c(1, 1)) &
  theme(plot.margin = margin(10, 12, 10, 18))

cutplt
```


```{r save-figure-5, echo=FALSE, warning=FALSE, message=FALSE}
ggsave(
  "figures/cutoff_sensi.png",
  plot = cutplt,
  width = 9,
  height = 8,
  dpi = 600,
  bg = "white"
)
ggsave(
  "figures/cutoff_sensi.pdf",
  plot = cutplt,
  width = 9,
  height = 8,
  bg = "white"
)
```



```{r, recompute person-level summary for the chosen cutoff}

cutoff_val <- best_cutoff$cutoff
  #round(cutoff_est, 5)
head(person_level_sed_all)
## This cut-off is not in the data, so we need to recompute the person-level summary for this specific cutoff
out_new <- compute_sleep_sed_stats_fixed(sleep_sed.df, cutoffs = cutoff_val, inc_slp_df = FALSE) 
## Get sedentary and sleep summaries for this cutoff
## Sedentary summary
new_person_level_sed <- person_level_sed.fun(df = out_new$sed_summ_df, cut = cutoff_val)

## Sleep summary
#new_person_level_sleep <- person_level_sleep.fun(df = out_new$sleep_summ_df, cut = cutoff_val)
#head(new_person_level_sleep)

## Add the column mean_sleep_time to the sedentary summary for merging
#new_person_level_sed_sleep <- new_person_level_sed %>%
#  left_join(
#    new_person_level_sleep %>% select(ID, mean_sleep_time),
#    by = "ID"
 # )


head(new_person_level_sed)

new_person_level_demo <- new_person_level_sed %>%
  left_join(
    demo_unique %>% 
      select(ID, Age_group, BMI_cat, Race, Gender,
             Education, Marital_status,
             Exam_weight_recal, PSU, Stratum),
    by = "ID"
  )
head(new_person_level_demo)
```


```{r Survey design at the PERSON level}

library(survey)

person_cut <- new_person_level_demo
  #subset(person_level_demo, cutoff == cutoff_val)

des_cut <- svydesign(
  id     = ~PSU,
  strata = ~Stratum,
  weights = ~Exam_weight_recal,
  nest   = TRUE,
  data   = person_cut
)

tail(new_person_level_demo)

length(unique(person_cut$ID))
```








## How sedentary are people with sleep disorder


```{r, compare sedentary time for people with and without sleep disorder, echo=FALSE, warning=FALSE}
head(person_cut)
head(self_reported_f.df)
ids <- self_reported_f.df$ID
slp_disorder.df <- person_cut %>% filter(ID %in% ids)
nrow(slp_disorder.df)


slp_disorder_new.df<- slp_disorder.df %>% left_join(self_reported_f.df, by='ID')
head(slp_disorder_new.df)
dim(slp_disorder_new.df)


## create a new design
des_new <- svydesign(
  id     = ~PSU,
  strata = ~Stratum,
  weights = ~Exam_weight_recal,
  nest   = TRUE,
  data   = slp_disorder_new.df
)
  sed_slptroble <- svyby(
    ~mean_sed_time,
    by = as.formula(paste("~", 'has_trouble_sleeping_cat')),
    design = des_new,
    FUN = svymean,
    vartype = c("se","ci"),
    na.rm = TRUE
  )
  sed_slptroble
  
  sed_slpdisorder <- svyby(
    ~mean_sed_time,
    by = as.formula(paste("~", 'has_sleep_disorder_cat')),
    design = des_new,
    FUN = svymean,
    vartype = c("se","ci"),
    na.rm = TRUE
  )
 sed_slpdisorder
 
## Compute percentage difference in sedentary time between people with and without sleep disorder
 
slp_disorder_yes <- coef(svymean(~mean_sed_time, subset(des_new, has_sleep_disorder_cat == "Yes")))
slp_disorder_no <- coef(svymean(~mean_sed_time, subset(des_new, has_sleep_disorder_cat == "No")))
perc_diff_slp_disorder <- ((slp_disorder_yes - slp_disorder_no) / slp_disorder_no) * 100
cat("Percentage difference in sedentary time between people with and without sleep disorder:", round(perc_diff_slp_disorder, 2), "%\n")


## Check table of sleep disorder categories to see how many people have sleep disorder and how many don't have sleep disorder in the sample
table(slp_disorder_new.df$has_sleep_disorder_cat)
910/(10512)
```


```{r, compare sleep time for people with and without sleep disorder, echo=FALSE, warning=FALSE}
table(slp_disorder_new.df$has_trouble_sleeping_cat)
table(slp_disorder_new.df$has_sleep_disorder_cat)

7.569495-6.722963
910/(10512)
9585+910+17
```


## Compare the estimate of sedentary time for the selected cutoff with others: Cutoff sensitivity analysis.


```{r, cutoff sensitivity analysis, echo=FALSE, warning=FALSE}

cutoff_range <- seq(30, 95, by = 5)
## include the selected cutoff in the range if it's not already there
if (!cutoff_est %in% cutoff_range) {
  cutoff_range <- sort(c(cutoff_range, cutoff_est))
}
head(cutoff_range)
cutoff_sensitivity <- compute_sleep_sed_stats_fixed(sleep_sed.df, cutoffs = cutoff_range)

# Apply function to all cutoffs and combine
new_person_level_sed_all <- bind_rows(
  lapply(cutoff_range, function(cut) person_level_sed.fun(df = cutoff_sensitivity$sed_summ_df, cut = cut))
)

new_person_level_sleep_all <- bind_rows(
  lapply(cutoff_range, function(cut) person_level_sleep.fun(df = cutoff_sensitivity$sleep_summ_df, cut = cut))
)

## Add the column mean_sleep_time to the sedentary summary for merging
new_person_level_sed_sleep_all <- new_person_level_sed_all %>%
  left_join(
    new_person_level_sleep_all %>% select(ID, mean_sleep_time, cutoff),
    by = c("ID", "cutoff")  # join by both ID and cutoff to ensure correct matching
  )
head(new_person_level_sed_sleep_all)
new1_person_level_demo <- new_person_level_sed_sleep_all %>%
  left_join(
    demo_unique %>% 
      select(ID, Age_group, BMI_cat, Race, Gender,
             Education, Marital_status,
             Exam_weight_recal, PSU, Stratum),
    by = "ID"
  )
head(new1_person_level_demo)
## Make a survey design for the sensitivity analysis data
des_sensitivity <- svydesign(
  id = ~PSU,
  strata = ~Stratum,
  weights = ~Exam_weight_recal,
  nest = TRUE,
  data = new1_person_level_demo
)
## Estimate sedentary time and 95% CI for each cutoff in the sensitivity analysis
sed_sensitivity_res <- svyby(
  ~mean_sed_time,
  by = ~cutoff,
  design = des_sensitivity,
  FUN = svymean,
  vartype = c("ci"),
  na.rm = TRUE
)

sed_sensitivity_res
## Make a coefficient plot to show how the sedentary time estimate and CI changes across cutoffs
plot_cutoff <- function(data,
                        x = "cutoff",
                        y = "mean_sed_time",
                        ci_l = "ci_l",
                        ci_u = "ci_u") {

  ggplot(data, aes(x = .data[[x]], y = .data[[y]])) +
    geom_point(size = 2) +
    geom_errorbar(aes(ymin = .data[[ci_l]], ymax = .data[[ci_u]]),
                  width = 1.5, linetype=1,
                  , col = "grey0") +
    theme_classic() +
    scale_x_continuous(breaks =round(data[[x]],1)) +
    scale_y_continuous(breaks=seq(min(round(data[[y]],2)),
                                max(round(data[[y]],2)+.5 ),.5)) +
    theme(
  axis.text.x = element_text(angle = 45, hjust = 1)
) +
    labs(
      x = "sleep percent cutoff",
      y = "Mean Sedentary Time (hours/day)",
    )
}

df <- sed_sensitivity_res
options(repr.plot.width = 10, repr.plot.height = 6)
p1 <- plot_cutoff(df)
p1 +
  geom_line(data = df, aes(x = cutoff, y = mean_sed_time),
            color = "grey", linewidth = 2) +   # thicker line

  geom_vline(xintercept = cutoff_est,
             linetype = "dashed",
             color = "grey",
             linewidth = 1.5) +                  # thicker vertical line

  annotate("text",
           x = cutoff_est,
           y = max(df$mean_sed_time, na.rm = TRUE),
           label = paste0(round(cutoff_est, 2), "%"),
           vjust = -0.5,
           color = "grey",
           fontface = "bold",
           size = 7) +                           # bigger annotation text

  theme(
    text = element_text(size = 16),              # global text size
    axis.title = element_text(size = 18),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16)
  )
```


```{r}
cutoff_table <- function(data, selected_cutoffs) {
  
  library(dplyr)
  
  data %>%
    filter(cutoff %in% selected_cutoffs) %>%
    mutate(
      estimate_ci = sprintf("%.3f (%.3f, %.3f)", mean_sed_time, ci_l, ci_u)
    ) %>%
    select(cutoff, estimate_ci)
}

# target <- round(cutoff_est,2)
cutoff_table(df, c(50,55,60,65,70,75,80,95,90,cutoff_est,95))
```


## Make plot of CHAP percentage sitting vs real sedentary time for the selected cutoff, stratified by day type (weekday vs weekend)


```{r, new function for plotting the hourly sedentary time with CIs, echo=FALSE, warning=FALSE}

new_plot_hourly_stratified_weighted_CI <- function(df, metric, group_var,
                                               title,
                                               show_legend = TRUE,
                                               legend_title = group_var) {
  
  library(survey)
  library(ggplot2)
  library(scales)
  library(dplyr)
  group_sym <- sym(group_var)

  # 1. Survey design using 2011–2014 pooled weight
  des <- svydesign(
    id     = ~PSU,
    strata = ~Stratum,
    weights = ~Exam_weight_recal,
    nest   = TRUE,
    data   = df
  )

  # 2. Weighted hourly means + CIs by group
  hourly_avg <- svyby(
    as.formula(paste0("~", metric)),
    by = as.formula(paste("~", group_var, "+ Hour")),
    design = des,
    FUN = svymean,
    na.rm = TRUE,
    vartype = c("ci")
  ) %>% 
    rename(avg_value = !!sym(metric))

  # Define manual linetypes and shapes
  groups <- unique(hourly_avg[[group_var]])
  n_groups <- length(groups)
  linetypes <- rep(c("twodash","solid","dotted","longdash", "dotdash",  "dashed"), length.out = n_groups)
  shapes    <- rep(c(16, 16, 16, 16, 8, 4), length.out = n_groups)

  # 3. Plot
  p <- hourly_avg %>%
    ggplot(aes(x = Hour, y = avg_value, group = !!group_sym)) +
    
    # CI ribbons (grey for all groups)
    geom_ribbon(aes(ymin = ci_l, ymax = ci_u, group = !!group_sym),
                fill = "grey80", alpha = 0.9, color = NA) +
    
    # Lines with different line types
    geom_line(aes(linetype = !!group_sym), color = "black", size = 2.5) +
    
    # Points with different shapes
    geom_point(aes(shape = !!group_sym), color = "black", size = 3) +
    
    # Formatting
    scale_y_continuous(
      limits = c(30, 90),
      breaks = seq(30, 90, 10),
      labels = percent_format(scale = 1)
    ) +
    scale_x_continuous(breaks = seq(1, 24, 2)) +
    
    # Apply manual linetypes and shapes
    scale_linetype_manual(values = linetypes) +
    scale_shape_manual(values = shapes) +
    
    # Labels
    labs(
      x = "Hour of Day",
      y = paste0("% ", gsub("_", " ", gsub("^percent_?", "", title))),
      linetype = legend_title,
      shape = legend_title
    ) +
    
    theme_classic(base_size = 30)

  # Toggle legend
  if (!show_legend) {
    p <- p + theme(legend.position = "none")
  }

  return(p)
}
```






```{r}
p1 <- new_plot_hourly_stratified_weighted_CI(sl_sed.df, metric="percent_sedentary_wrist", group_var="DayType",
                                         title = "Sitting",
                                         show_legend = FALSE)
                                        #legend_title = group_var)

new_sl_sed.df <- sl_sed.df %>% filter(percent_sleep <
   cutoff_est)

head(new_sl_sed.df)

## find how many unique IDs are in each Hour for the new_sl_sed.df
unique_ids_per_hour <- new_sl_sed.df %>%
  group_by(Hour) %>%
  summarise(n_unique_IDs = length(unique(ID)))

length(unique(new_sl_sed.df$ID))

p2 <- new_plot_hourly_stratified_weighted_CI(new_sl_sed.df, metric="percent_sedentary_wrist", group_var="DayType", title = "sedentary",
                                         show_legend = FALSE)

 
(p1 + p2) +
  plot_layout(
    guides = "collect"    # one shared legend
  ) 

# &
#   theme(
#     legend.position = "right"   # or "bottom", "top", etc.
#   )

head(sl_sed.df)

plot(unique_ids_per_hour$Hour, unique_ids_per_hour$n_unique_IDs, type = "b",
     xlab = "Hour", ylab = "Number of Unique IDs",
     main = "Unique IDs per Hour in new_sl_sed.df")
```


```{r}
## Let stratify by BMI category
## make BMI category variable
sl_sed.df <- sl_sed.df %>%
  mutate( BMI=Weight / (Height/100)^2) %>%
  mutate(
    BMI_cat = case_when(
    BMI < 25 ~ "under/normal weight",
    BMI >= 25 & BMI < 30 ~ "over weight",
    BMI >= 30 ~ "obesity",
    TRUE ~ NA_character_
  ))
  

p3 <- new_plot_hourly_stratified_weighted_CI(sl_sed.df, metric="percent_sedentary_wrist", group_var="BMI_cat",
                                         title = "Sitting",
                                         show_legend = FALSE)
                                         #,legend_title = "BMI Category")
p3

head(sl_sed.df)
new_sl_sed.df <- sl_sed.df %>% filter(percent_sleep < cutoff_est)
p4 <- new_plot_hourly_stratified_weighted_CI(new_sl_sed.df, metric="percent_sedentary_wrist", group_var="BMI_cat",
                                         title = "Sedentary",
                                         show_legend = FALSE)
                                         #legend_title = "BMI Category")

(p3 + p4) +
  plot_layout(
    guides = "collect"    # one shared legend
  )


```






















## Explore functional analysis (sedentary and sleep data separately)

```{r Sedentary data: functional analysis, echo=FALSE, warning=FALSE}
head(sl_sed.df)

funcdat_format <- function(df, coln ='percent_sleep'){
  
  #cnames <- colnames(df)
  if(coln=='percent_sedentary'){
    df1 <- df %>% select(ID,Day,Hour,percent_sedentary)
    #head(sedfunanl.df)
    ## Summarise into 24 hours data per participants
    df1new <- df1 %>%
      summarise(
        avg_sed = mean(percent_sedentary/100, na.rm = TRUE),
        .by = c(ID, Hour)
      )
    long.df <- df1new %>% pivot_wider(names_from = 'Hour', values_from = 'avg_sed',
                                      names_prefix = 'Hour_')
    
  }else{
    df1 <- df %>% select(ID,Day,Hour,percent_sleep)
    ## Summarise into 24 hours data per participants
    df1new <- df1 %>%
      summarise(
        avg_sleep = mean(percent_sleep/100, na.rm = TRUE),
        .by = c(ID, Hour)
      )
    long.df <- df1new %>% pivot_wider(names_from = 'Hour', values_from = 'avg_sleep',
                                      names_prefix = 'Hour_')
  }
  return(long.df)
}


```

```{r Reshape data for function anaysis: Sedentary data, echo=FALSE, warning=FALSE}

## Sedentary data reshape in a wider formate
sedfunanl.df <- funcdat_format(df= sl_sed.df, coln = 'percent_sedentary')

sleepfunanl.df <- funcdat_format(df= sl_sed.df)

head(sedfunanl.df)

write.csv(sedfunanl.df, file = 'outputs/sedFDA.csv')
head(sleepfunanl.df)
nrow((sleepfunanl.df))
```

## Histogram for sedentary & sleep hours

```{r histogram of sedentary hours}
hists <- function(df, coln='sleep'){
df2 <- apply(
  df[ , -1],   # all hour columns, remove ID
  1,                         # row-wise
  FUN = function(x) mean(x, na.rm = TRUE)
)

df2 <- data.frame(
  ID = df$ID,
  avg_row_mean = df2
)
xlab <- ifelse(coln=='sed','sedentary per hour','sleep per hour') 
return(hist(df2$avg_row_mean , nclass=10, main = NULL, xlab = xlab ,probability = T))
}

par(mfrow=c(1,2))
# histogram for sleep hours
hists(df= sleepfunanl.df)
## histogram for sedentary hours
hists(df=sedfunanl.df, coln = 'sed')

```

## Run PCA on sedentary and sleep data:
- The goal is to revel the main patterns of variation in daily sedentary and sleep behaviour.

```{r}

spca <- function(df){
## remove ID and turn into a matrix
Mat <- as.matrix(df[,-1])
## Run PCA
pca <- prcomp(Mat, scale. = TRUE)
return(pca)
}

## run pca for sleep
sleep_pca <- spca(df=sleepfunanl.df)
## run pca for sedentary
sed_pca <- spca(df=sedfunanl.df)

print("=======Summarising sleep pca output===========")
summary(sleep_pca)
print("===End of sleep pca summary")
print("====***********=====")

print("====Summarise sedentary pca results=======")
summary(sed_pca)
```

```{r plot Scree plot for pca results}
par(mfrow=c(1,2))
plot(sed_pca, type = "l", main = "Sed Scree Plot")
plot(sleep_pca, type = "l", main = "Sleep Scree Plot")

```
## PCA scores tells the individual participants values for each pca (linear combinations of the original variables)

```{r Extract PCA scores}

extr_scores <- function(pca.df,df=sleepfunanl.df){
scores <- as.data.frame(pca.df$x)

scores<-scores %>%  mutate(ID=df$ID )

return(scores)
}

sed_scores <- extr_scores(pca.df = sed_pca,
                          df=sedfunanl.df)
sleep_scores <- extr_scores(pca.df = sleep_pca)
head(sed_scores)
head(sleep_scores)
```

```{r plot pca scores}

scoreplot <-  function(scores, coln='sleep'){
 # Subset data
dat <- scores[1:100, ]

# Base scatter plot
main = ifelse(coln=='sed', "Sedentary PCA scores",  "Sleep PCA scores")
plot(dat$PC1, dat$PC2,
     main = main,
     xlab = "PC1",
     ylab = "PC2",
     pch = 19,
     col = "blue")

# Add text labels
text(dat$PC1, dat$PC2,
     labels = dat$ID,
     pos = 3,   # 3 = above point
     cex = 0.7)
}

par(mfrow=c(1,2))
scoreplot(scores = sleep_scores)
scoreplot(scores = sed_scores, coln='sed')

```


## Get Hour Loadings(Coefficient of the linear combination of original 24 hours variables) from PCA: Which hours matters the most?

```{r extract loadings}
extrloadings <- function(pca.df){

loadings <- as.data.frame(pca.df$rotation)
loadings$Hour <- rownames(loadings)

return(loadings)
}

sleep_loadings <- extrloadings(sleep_pca)
sed_loadings <- extrloadings(sed_pca)
head(sleep_loadings)
```

```{r Define a plot function for loadings}

plotloadings <- function(loadings, coln='sleep'){
# Convert loadings to long format for ggplot
load_long <- loadings %>%
  pivot_longer(
    cols = starts_with("PC"),
    names_to = "PC",
    values_to = "loading"
  ) %>%
  filter(PC %in% c("PC1", "PC2", "PC3", "PC4"))

## Make hour ordered factor
load_long$Hour <- factor(
  load_long$Hour,
  levels = paste0("Hour_", 1:24),
  ordered = TRUE
)
title <- ifelse(coln=='sed',"Sedentary PC1–PC4 loadings","Sleep PC1–PC4 loadings" )
# Plot
ggplot(load_long, aes(x = Hour, y = loading, fill = PC)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ PC, ncol = 1, scales = "free_y") +
  labs(
    title = title,
    x = "Hour",
    y = "Loading Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
}


```

## plot loadings

```{r plot loadings and get interpretations: sleep loadings}
pload_sleep <- plotloadings(loadings = sleep_loadings)
pload_sleep
```

```{r plot loadings and get interpretations: sedentary loadings}

pload_sed <- plotloadings(loadings = sed_loadings, coln = 'sed')
(pload_sed + pload_sleep)
```

## Possible meaning of the PCA's (PC1-PC2)

```{r}
pc_summary <- data.frame(
  PC = paste0("PC", 1:4),
  Interpretation = c(
    "Overall sedentary intensity (high = more sedentary across whole day)",
    "Morning vs midday pattern (high = morning sedentary, low = midday sedentary)",
    "Midday peak vs midday active pattern (high = midday sitting peak)",
    "Subtle timing variation (early vs late minor differences)"
  )
)

pc_summary

```



## K-means clustering using pca to see hiding cluster of sleep or sedentary time

```{r}

kmeanplot <- function(scores,k, seed, coln='sleep'){
# K = 2, 3, or 4 depending on your dataset size
set.seed(seed)

km <- kmeans(scores[, c("PC1","PC2","PC3")], centers = k)
scores$cluster <- factor(km$cluster)

# Visualize clusters
title <-  ifelse( coln =='sed',"Sedentary Clusters PC1-PC2", "Sleep Clusters PC1-PC2" )
p1= ggplot(scores, aes(PC1, PC2, color = cluster, label = ID)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.1, size = 3) +
  labs(title = title) +
  theme_minimal()

return(list(scores=scores,plt=p1 ))
} 
```




```{r sleep & sedentary k means cluster}
sleep_clusters <- kmeanplot(scores = sleep_scores, k=4, seed = 23 )
sleep_cluster <- sleep_clusters$scores
sleep_cluster_plot <- sleep_clusters$plt


sed_clusters <- kmeanplot(scores = sed_scores, k=4, seed = 23, coln = 'sleep')
sed_cluster <- sed_clusters$scores
sed_cluster_plot <- sed_clusters$plt


sed_cluster_plot+sleep_cluster_plot


```



## Cluster specific mean curves

```{r mean specific clusters}

cluster_mean_curves <- function(scores_cluster, df, coln='sleep', use_other_cl="No"){
  # Allow to use sedentary clusters to get their sleep curves
# Merge cluster labels back to the wide data
scores <- scores_cluster
if(use_other_cl=="No"){
wide_clustered <- cbind(scores["cluster"], df)
}
else{
  
  wide_clustered <- cbind(scores["other_cluster"], df)
  names(wide_clustered)[names(wide_clustered) == "other_cluster"] <- "cluster"

}
# pivot longer for plotting
long_clustered <- wide_clustered %>%
  pivot_longer(
    cols = starts_with("Hour_"),
    names_to = "Hour",
    values_to = "sed"
  )
long_clustered$Hour <- factor(
  long_clustered$Hour,
  levels = paste0("Hour_", 1:24),
  ordered = TRUE
)

cluster_sizes <- long_clustered %>%
  distinct(ID, cluster) %>%       # make sure each ID counts once
  count(cluster, name = "n_participants")



# Remove ID column that got carried in
long_clustered$ID <- NULL

# Compute mean curves
cluster_means <- long_clustered %>%
  group_by(cluster, Hour) %>%
  summarise(mean_sed = mean(sed, na.rm = TRUE), .groups = "drop")

cluster_means <- cluster_means %>%
  left_join(cluster_sizes, by = "cluster")


# ---- Define color palettes based on coln ----
n_clust <- length(unique(cluster_means$cluster))

sleep_palette <- scales::hue_pal(h = c(210, 300), c = 80, l = 50)(n_clust)
# sed_palette <- scales::hue_pal(
#   h = c(220, 20),   # blue → red
#   c = 80,
#   l = 50
# )(n_clust)
#sed_palette <- viridisLite::plasma(n_clust)
sed_palette <- viridisLite::cividis(n_clust)
# sed_palette   <- viridisLite::inferno(n_clust)



cluster_colors <- if (coln == "sleep") sleep_palette else sed_palette
names(cluster_colors) <- sort(unique(cluster_means$cluster))


# Plot cluster-specific mean curves
#title<- ifelse(coln=='sed', "24-Hour Sedentary Curves",
            #   "24-Hour Sleep Curves")
yla<- ifelse(coln=='sed', "Mean Sedentary %",
              "Mean Sleep %")
cluster_means$Hour_num <- as.numeric(sub("Hour_", "", cluster_means$Hour))
p1<- ggplot(cluster_means, aes(x = Hour_num, y = mean_sed, color = cluster, group = cluster)) +
  geom_line(size = 1.3) +
  geom_point(size = 2) +
  scale_color_manual(values = cluster_colors, name = "Cluster") +
  labs(
    title = NULL,
    x = "Hour",
    y = yla
  ) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Dynamic titles based on coln
bar_title <- ifelse(coln == "sed",
                    "Cluster Sizes (Sedentary Analysis)",
                    "Cluster Sizes (Sleep Analysis)")

bar_y_label <- ifelse(coln == "sed",
                      "Number of Participants (Sedentary)",
                      "Number of Participants (Sleep)")

p_bar <- ggplot(cluster_sizes,
                aes(x = factor(cluster),
                    y = n_participants,
                    fill = factor(cluster))) +
  geom_col(width = 0.65) +
  geom_text(aes(label = n_participants),
            vjust = -0.5,
            size = 4) +
  scale_fill_manual(values = cluster_colors) +
  labs(
    title =NULL, # bar_title,
    x = "Cluster",
    y = bar_y_label
  ) +
  scale_y_continuous(
    limits = c(0,10000)
              )+
  theme_classic(base_size = 14) +
  theme(legend.position = "none")

return(list(
  mean_curve_plot = p1,
  cluster_bar_plot = p_bar
))

}

head(sleep_cluster)
```




```{r plot cluster specific mean curves}
## Add sedentary clusters to sleep time data
sleep_cluster$other_cluster <- sed_cluster$cluster

## Add sleep clusters to sedentary time data
sed_cluster$other_cluster <- sleep_cluster$cluster


## Use original clusters
sleep_curves_orig <- cluster_mean_curves(scores_cluster=sleep_cluster , df=sleepfunanl.df, coln='sleep')

sed_curves_orig <- cluster_mean_curves(scores_cluster=sed_cluster , df=sedfunanl.df, coln='sed')

## Switch clusters
sleep_curves_switch <- cluster_mean_curves(scores_cluster=sleep_cluster , df=sleepfunanl.df, coln='sleep', use_other_cl = "Yes")

sed_curves_switch <- cluster_mean_curves(scores_cluster=sed_cluster , df=sedfunanl.df, coln='sed', use_other_cl = "Yes")
```


```{r sleep mean and bar plot with original clusters}
## Mean curves
sleep_curves_orig$cluster_bar_plot+sleep_curves_orig$mean_curve_plot
```

```{r sedentary mean and bar plot with original clusters}
## Sed Mean curves and total participants in cluster
(sleep_curves_orig$cluster_bar_plot+sleep_curves_orig$mean_curve_plot)/(sed_curves_orig$cluster_bar_plot+sed_curves_orig$mean_curve_plot)
```





```{r sleep mean and bar plot with with sedentary clusters}
## Mean curves
sleep_curves_switch$cluster_bar_plot+sleep_curves_switch$mean_curve_plot
```


```{r sedentary mean and bar plot with sleep clusters}
## Sed Mean curves and total participants in cluster
sed_curves_switch$cluster_bar_plot+sed_curves_switch$mean_curve_plot
```


```{r sleep clusters and their corresponding sed curves}
sleep_curves_orig$mean_curve_plot+sed_curves_switch$mean_curve_plot
```

```{r sedentary clusters and their corresponding sleep curves}
(sed_curves_orig$mean_curve_plot+ sleep_curves_switch$mean_curve_plot)/(sleep_curves_orig$mean_curve_plot+sed_curves_switch$mean_curve_plot)
```

## Get the sleep aware sedentary time for these groups (clusters)

```{r}
sed_clusters <- sed_cluster %>% select(ID, cluster)
head(sed_clusters)
nrow(sed_clusters)

derived_sed_cluster <- person_cut %>% 
  left_join(sed_clusters, by="ID")
head(derived_sed_cluster)
nrow(derived_sed_cluster)
```


```{r}

## create a new design
newdes_cut  <- svydesign(
  id     = ~PSU,
  strata = ~Stratum,
  weights = ~Exam_weight_recal,
  nest   = TRUE,
  data   = derived_sed_cluster
)
tab_cluster <- make_subgroup_table(
  design = newdes_cut,
  var = "cluster",
  #levels = c("under/normal weight", "over weight", "obesity"),
  strat_var = "Age_group",
  strat_levels = c("<18","18-34","35-44","45-64", ">=65"),
  include_overall = TRUE
)
tab_cluster
#derived_sed_cluster
```



## Other analysis

```{r}
pca_pattern_plot <- function(scores,
                             pc1 = "PC1",
                             pc2 = "PC2",
                             pc3 = "PC3",
                             coln = c("sed", "sleep"),
                             thr_PC2 = 1,
                             thr_PC3 = 1) {
  
  # Validate column names
  if (!all(c(pc1, pc2, pc3) %in% colnames(scores))) {
    stop("One or more PC columns not found in scores dataframe.")
  }
  
  # Validate type
  coln <- match.arg(coln)
  
  # Extract values dynamically
  PC1_vals <- scores[[pc1]]
  PC2_vals <- scores[[pc2]]
  PC3_vals <- scores[[pc3]]
  
  # ---- 1. Assign pattern labels generically ----
  scores$pattern <- ifelse(
    PC2_vals > thr_PC2,  "Morning-heavy",
    ifelse(PC2_vals < -thr_PC2, "Evening-heavy",
      ifelse(PC3_vals > thr_PC3, "Midday-peak",
             "Flat/Uniform")
    )
  )
  
  # ---- 2. Pattern counts ----
  pattern_counts <- table(scores$pattern)
  
  # ---- 3. Dynamic title ----
  pattern_title <- ifelse(
    coln == "sed",
    "Functional Sedentary Pattern Types (PCA-based)",
    "Functional Sleep Pattern Types (PCA-based)"
  )
  
  # ---- 4. PCA scatter plot ----
  p_pattern <- ggplot(scores, aes_string(x = pc1, y = pc2, color = "pattern")) +
    geom_point(size = 3) +
    labs(
      title = pattern_title,
      x = pc1,
      y = pc2,
      color = "Pattern"
    ) +
    theme_minimal()
  
  # ---- 5. Return everything ----
  return(list(
    plot = p_pattern,
    counts = pattern_counts,
    scores = scores
  ))
}


```


```{r}
pca_pattern_plot(sed_scores, coln = "sed")

```


```{r}
out_sleep <- pca_pattern_plot(sleep_scores,
                              coln = "sleep")
out_sleep
```


```{r sleep & sedentary intensity using PC1}

make_intensity_index <- function(scores,
                                 pc_col = "PC1",
                                 coln = c("sed", "sleep"),
                                 scale_to = c(0, 100)) {
  
  # Validate column name
  if (!pc_col %in% colnames(scores)) {
    stop(paste0("Column '", pc_col, "' not found in scores dataframe"))
  }
  
  # Validate type
  coln <- match.arg(coln)
  
  # ---- 1. Create index generically ----
  index_vals <- scales::rescale(scores[[pc_col]], to = scale_to)
  
  # Dynamic name: sed_intensity_index or sleep_intensity_index
  index_name <- ifelse(
    coln == "sed",
    "sedentary_intensity_index",
    "sleep_intensity_index"
  )
  
  # Attach to scores
  scores[[index_name]] <- index_vals
  
  # ---- 2. Dynamic plot title ----
  plot_title <- ifelse(
    coln == "sed",
    "Sedentary Intensity Index (0–100)",
    "Sleep Intensity Index (0–100)"
  )
  
  # ---- 3. Histogram ----
  p_hist <- ggplot(scores, aes_string(index_name)) +
    geom_histogram(bins = 20, fill = "steelblue") +
    labs(
      title = plot_title,
      x = index_name,
      y = "Count"
    ) +
    theme_minimal()
  
  # ---- 4. Return everything ----
  return(list(
    plot = p_hist,
    scores = scores,
    index_name = index_name
  ))
}


```


```{r}
out_sed <- make_intensity_index(sed_scores,
                                pc_col = "PC1",
                                coln = "sed")

out_sed$plot          # histogram
head(out_sed$scores)  # includes sedentary_intensity_index

```
```{r}
out_sleep <- make_intensity_index(sleep_scores,
                                  pc_col = "PC1",
                                  coln = "sleep")

out_sleep$plot
head(out_sleep$scores)

```
