---
title: "Final Sedentary-Time Analysis"
author: "Christopher Odoom"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    toc: true
    toc_depth: 2
    fig_caption: true
header-includes:
  - \usepackage{booktabs}
---

```{r setup, include=FALSE}
find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, winslash = "/", mustWork = TRUE)
  repeat {
    if (dir.exists(file.path(current, "reports")) && dir.exists(file.path(current, "R"))) {
      return(current)
    }
    parent <- dirname(current)
    if (identical(parent, current)) stop("Could not find project root.")
    current <- parent
  }
}

project_root <- find_project_root()
knitr::opts_knit$set(root.dir = project_root)
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.extra = ""
)
options(knitr.kable.NA = "")

output_root <- file.path(project_root, "outputs", "final_analysis")
figure_root <- file.path(project_root, "figures", "final_analysis")
output_dirs <- file.path(
  output_root,
  c("tables", "results", "cutoff", "qc", "datasets", "documentation")
)
invisible(lapply(c(output_root, output_dirs, figure_root), dir.create, recursive = TRUE, showWarnings = FALSE))
```

## Purpose

This R Markdown file documents and reproduces the manuscript analysis for
estimating waking sedentary time in NHANES 2011-2014 wrist accelerometer data.
The project combines posture predictions from the CHAP wrist model with
sleep/nonwear information from SWaN to estimate each participant's usual daily
waking sedentary time.

The main goal is to produce survey-weighted, nationally representative estimates
of waking sedentary time after excluding sleep/nonwear periods. The analysis
calibrates a sleep/nonwear percentage cutoff against NHANES self-reported sleep
duration, applies the selected cutoff at the primary analysis epoch, summarizes
sedentary time to the participant level, and estimates overall and subgroup means
using the NHANES survey design.

This file is also a reproducibility driver. It reads the analysis-ready
accelerometer datasets from the `wristsed` data package, derives the analytic
sample, recalibrates the sleep cutoff for the 1-hour SWaN primary analysis,
generates manuscript tables and figures, and writes final outputs to
`outputs/final_analysis/` and `figures/final_analysis/`.

## Packages

```{r packages}
library(dplyr)
library(ggplot2)
library(haven)
library(knitr)
library(patchwork)
library(purrr)
library(readr)
library(scales)
library(survey)
library(tidyr)
library(wristsed)

options(survey.lonely.psu = "adjust")
```

## Analysis Configuration

The primary render uses the 1-hour SWaN epoch dataset. Sleep/nonwear removal is
applied at the epoch level before participant-day sedentary totals are
calculated. For figures that display hourly patterns, records are aggregated to
hour only after the sleep/nonwear rule has been applied.

The primary analytic unit is the NHANES participant. Epoch-level accelerometer
records are used to construct participant-day and participant-level sedentary
time summaries, but survey-weighted estimation is performed at the participant
level to avoid giving participants with more observed records or more valid days
disproportionate influence.

## Data Access

```{r data-access}
read_participant <- function(sleep_method = "swan", epoch = "1h") {
  get_participant_data(
    sleep_method = sleep_method,
    epoch = epoch,
    collect = TRUE
  )
}

read_epoch <- function(sleep_method = "swan", epoch = "1h") {
  get_epoch_data(
    sleep_method = sleep_method,
    epoch = epoch,
    collect = TRUE
  )
}

assert_complete_survey_data <- function(data, label) {
  missing_survey <- data %>%
    filter(is.na(Exam_weight_recal) | is.na(PSU) | is.na(Stratum))

  if (nrow(missing_survey) > 0L) {
    stop(
      label,
      " has participants without complete survey metadata: ",
      paste(missing_survey$ID, collapse = ", "),
      call. = FALSE
    )
  }

  data
}

primary_package_participant <- read_participant("swan", "1h") %>%
  assert_complete_survey_data("SWaN participant data")
nhanes_package_participant <- read_participant("nhanes", "1h") %>%
  assert_complete_survey_data("NHANES participant data")
primary_epoch <- read_epoch("swan", "1h")

metadata_columns <- c(
  "ID", "Dataset", "Age_group", "BMI_cat", "Race", "Gender", "Education",
  "Marital_status", "Exam_weight_recal", "PSU", "Stratum"
)

participant_metadata <- bind_rows(
  primary_package_participant,
  nhanes_package_participant
) %>%
  select(any_of(metadata_columns)) %>%
  distinct(ID, Dataset, .keep_all = TRUE)

participant_weights <- participant_metadata %>%
  distinct(ID, Dataset, Exam_weight_recal, PSU, Stratum)
```

## Reusable Functions

```{r reusable-functions}
format_decimal <- function(x, digits = 4) {
  formatC(as.numeric(x), format = "f", digits = digits)
}

format_mean_se <- function(mean, se, digits = 2) {
  paste0(format_decimal(mean, digits), " (", format_decimal(se, digits), ")")
}

format_render_table <- function(df, digits = 2) {
  df %>%
    mutate(across(where(is.numeric), ~round(.x, digits)))
}

paper_kable <- function(df, caption, digits = 2, align = NULL, latex_size = "\\small") {
  rendered <- format_render_table(df, digits = digits)

  if (knitr::is_latex_output()) {
    table_text <- knitr::kable(
      rendered,
      format = "latex",
      booktabs = TRUE,
      linesep = "",
      caption = caption,
      align = align
    )
    knitr::asis_output(paste(latex_size, table_text, "\\normalsize", sep = "\n"))
  } else {
    html_align <- if (!is.null(align) && any(grepl("p\\{", align))) NULL else align
    knitr::kable(rendered, caption = caption, align = html_align)
  }
}

format_export_numbers <- function(df) {
  df %>%
    mutate(across(where(is.numeric), ~round(.x, 4)))
}

save_final_figure <- function(plot, filename, width, height) {
  png_path <- file.path(figure_root, paste0(filename, ".png"))
  pdf_path <- file.path(figure_root, paste0(filename, ".pdf"))

  ggsave(png_path, plot, width = width, height = height, units = "in", dpi = 300, bg = "white")
  ggsave(pdf_path, plot, width = width, height = height, units = "in", bg = "white")

  invisible(c(png_path, pdf_path))
}

svy_design <- function(data) {
  svydesign(
    id = ~PSU,
    strata = ~Stratum,
    weights = ~Exam_weight_recal,
    nest = TRUE,
    data = data
  )
}

survey_mean_row <- function(design, variable, label) {
  estimate <- svymean(as.formula(paste0("~", variable)), design, na.rm = TRUE)
  ci <- confint(estimate)
  tibble(
    Measure = label,
    `Mean (h/day)` = unname(coef(estimate)[1]),
    SE = unname(SE(estimate)[1]),
    `95% CI lower` = unname(ci[1, 1]),
    `95% CI upper` = unname(ci[1, 2])
  )
}

safe_format_mean_ci <- function(design_obj, variable = "est_sedentary_hours", digits = 2) {
  values <- design_obj$variables[[variable]]
  if (nrow(design_obj$variables) == 0L || !any(!is.na(values))) {
    return(NA_character_)
  }

  tryCatch(
    {
      estimate <- svymean(as.formula(paste0("~", variable)), design_obj, na.rm = TRUE)
      ci <- confint(estimate)
      sprintf(
        paste0("%.", digits, "f (%.", digits, "f, %.", digits, "f)"),
        coef(estimate)[1],
        ci[1, 1],
        ci[1, 2]
      )
    },
    error = function(e) NA_character_
  )
}

format_n_pct <- function(df, denominator, digits = 1) {
  n <- n_distinct(df$ID)
  paste0(comma(n), " (", format_decimal(100 * n / denominator, digits), "%)")
}

valid_day_source_epochs <- function(sleep_method) {
  switch(sleep_method, swan = 8640L, nhanes = 1440L)
}

valid_epoch_days <- function(epoch_df, sleep_method) {
  epoch_df %>%
    group_by(ID, Dataset, Day) %>%
    summarise(day_source_epochs = sum(n_source_epochs, na.rm = TRUE), .groups = "drop") %>%
    filter(day_source_epochs >= valid_day_source_epochs(sleep_method)) %>%
    select(ID, Dataset, Day)
}

participant_from_epoch_cutoff <- function(epoch_df, cutoff, sleep_method) {
  valid_days <- valid_epoch_days(epoch_df, sleep_method)

  day_summary <- epoch_df %>%
    inner_join(valid_days, by = c("ID", "Dataset", "Day")) %>%
    mutate(
      duration_hours = epoch_duration_minutes / 60,
      waking_epoch = percent_sleep_nonwear < cutoff,
      sed_hours = if_else(waking_epoch, percent_sedentary / 100 * duration_hours, 0),
      wake_hours = if_else(waking_epoch, duration_hours, 0),
      sleep_nonwear_hours = if_else(waking_epoch, 0, duration_hours)
    ) %>%
    group_by(ID, Dataset, Day, DayType) %>%
    summarise(
      sedentary_hours = sum(sed_hours, na.rm = TRUE),
      waking_hours = sum(wake_hours, na.rm = TRUE),
      sleep_nonwear_hours = sum(sleep_nonwear_hours, na.rm = TRUE),
      .groups = "drop"
    )

  day_summary %>%
    group_by(ID, Dataset) %>%
    summarise(
      sleep_nonwear_cutoff = cutoff,
      no_of_valid_days = n(),
      est_sedentary_hours = mean(sedentary_hours, na.rm = TRUE),
      sd_est_sedentary_hours = sd(sedentary_hours, na.rm = TRUE),
      se_est_sedentary_hours = sd_est_sedentary_hours / sqrt(no_of_valid_days),
      est_waking_hours = mean(waking_hours, na.rm = TRUE),
      sd_waking_hours = sd(waking_hours, na.rm = TRUE),
      se_waking_hours = sd_waking_hours / sqrt(no_of_valid_days),
      est_sleep_nonwear_hours = mean(sleep_nonwear_hours, na.rm = TRUE),
      sd_sleep_nonwear_hours = sd(sleep_nonwear_hours, na.rm = TRUE),
      se_sleep_nonwear_hours = sd_sleep_nonwear_hours / sqrt(no_of_valid_days),
      .groups = "drop"
    ) %>%
    left_join(participant_metadata, by = c("ID", "Dataset")) %>%
    filter(!is.na(Exam_weight_recal), !is.na(PSU), !is.na(Stratum))
}

participant_from_direct_state <- function(epoch_df, sleep_method) {
  valid_days <- valid_epoch_days(epoch_df, sleep_method)

  epoch_df %>%
    inner_join(valid_days, by = c("ID", "Dataset", "Day")) %>%
    mutate(
      duration_hours = epoch_duration_minutes / 60,
      sedentary_hours = percent_waking_sedentary / 100 * duration_hours,
      waking_hours = percent_wake / 100 * duration_hours,
      sleep_nonwear_hours = percent_sleep_nonwear / 100 * duration_hours
    ) %>%
    group_by(ID, Dataset, Day, DayType) %>%
    summarise(
      sedentary_hours = sum(sedentary_hours, na.rm = TRUE),
      waking_hours = sum(waking_hours, na.rm = TRUE),
      sleep_nonwear_hours = sum(sleep_nonwear_hours, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    group_by(ID, Dataset) %>%
    summarise(
      no_of_valid_days = n(),
      est_sedentary_hours = mean(sedentary_hours, na.rm = TRUE),
      est_waking_hours = mean(waking_hours, na.rm = TRUE),
      est_sleep_nonwear_hours = mean(sleep_nonwear_hours, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(participant_weights, by = c("ID", "Dataset")) %>%
    filter(!is.na(Exam_weight_recal), !is.na(PSU), !is.na(Stratum))
}

estimate_three_means <- function(data) {
  design <- svy_design(data)
  sed_est <- svymean(~est_sedentary_hours, design, na.rm = TRUE)
  wake_est <- svymean(~est_waking_hours, design, na.rm = TRUE)
  sleep_est <- svymean(~est_sleep_nonwear_hours, design, na.rm = TRUE)

  tibble(
    mean_sedentary_hours = unname(coef(sed_est)[1]),
    se_sedentary_hours = unname(SE(sed_est)[1]),
    mean_waking_hours = unname(coef(wake_est)[1]),
    se_waking_hours = unname(SE(wake_est)[1]),
    mean_sleep_nonwear_hours = unname(coef(sleep_est)[1]),
    se_sleep_nonwear_hours = unname(SE(sleep_est)[1])
  )
}

estimate_cutoff_curve_row <- function(data, cutoff) {
  design <- svy_design(data)
  sleep_est <- svymean(~est_sleep_nonwear_hours, design, na.rm = TRUE)
  sed_est <- svymean(~est_sedentary_hours, design, na.rm = TRUE)

  tibble(
    cutoff = cutoff,
    sleep_mean = unname(coef(sleep_est)[1]),
    sleep_se = unname(SE(sleep_est)[1]),
    sed_mean = unname(coef(sed_est)[1]),
    sed_se = unname(SE(sed_est)[1])
  )
}

cutoff_grid <- function() {
  seq(50, 100, by = 0.01)
}

mean_sleep_nonwear_by_cutoff <- function(epoch_weights, cutoffs, total_person_weight) {
  keep <- is.finite(epoch_weights$percent_sleep_nonwear) &
    is.finite(epoch_weights$epoch_weight) &
    epoch_weights$epoch_weight > 0

  if (!any(keep)) {
    stop("No valid epoch weights are available for cutoff selection.")
  }

  percent_sleep <- epoch_weights$percent_sleep_nonwear[keep]
  weight <- epoch_weights$epoch_weight[keep]
  order_index <- order(percent_sleep)
  percent_sleep <- percent_sleep[order_index]
  weight <- weight[order_index]

  cumulative_weight <- cumsum(weight)
  total_epoch_weight <- sum(weight)
  lower_index <- findInterval(cutoffs, percent_sleep, left.open = TRUE)
  lower_weight <- ifelse(lower_index > 0L, cumulative_weight[lower_index], 0)
  (total_epoch_weight - lower_weight) / total_person_weight
}

cutoff_objective_curve <- function(epoch_df, sleep_method, self_sleep_mean) {
  cutoffs <- cutoff_grid()
  valid_days <- valid_epoch_days(epoch_df, sleep_method)

  person_days <- epoch_df %>%
    inner_join(valid_days, by = c("ID", "Dataset", "Day")) %>%
    distinct(ID, Dataset, Day) %>%
    count(ID, Dataset, name = "n_days")

  person_weights <- participant_weights %>%
    inner_join(person_days, by = c("ID", "Dataset")) %>%
    filter(!is.na(Exam_weight_recal), !is.na(n_days))

  epoch_weights <- epoch_df %>%
    inner_join(valid_days, by = c("ID", "Dataset", "Day")) %>%
    select(ID, Dataset, percent_sleep_nonwear, epoch_duration_minutes) %>%
    inner_join(person_weights, by = c("ID", "Dataset")) %>%
    transmute(
      percent_sleep_nonwear,
      epoch_weight = (Exam_weight_recal / n_days) * (epoch_duration_minutes / 60)
    )

  total_person_weight <- sum(person_weights$Exam_weight_recal, na.rm = TRUE)

  tibble(cutoff = cutoffs) %>%
    mutate(
      sleep_mean = mean_sleep_nonwear_by_cutoff(epoch_weights, cutoff, total_person_weight),
      diff = abs(sleep_mean - self_sleep_mean)
    )
}

select_cutoff_for_epoch <- function(epoch_df, sleep_method, self_sleep_mean) {
  objective_curve <- cutoff_objective_curve(epoch_df, sleep_method, self_sleep_mean)
  selected <- objective_curve %>%
    arrange(diff, cutoff) %>%
    slice(1)
  participant <- participant_from_epoch_cutoff(epoch_df, selected$cutoff, sleep_method)
  estimates <- estimate_three_means(participant)

  bind_cols(
    selected %>% select(cutoff, sleep_mean, diff),
    estimates
  )
}
```

## Self-Reported Sleep Benchmark

NHANES self-reported sleep duration is used as an external benchmark for
selecting the accelerometer sleep/nonwear cutoff. The benchmark is estimated
with the same combined 2011-2014 survey design used for the primary participant
analysis.

```{r self-reported-sleep}
self_sleep <- bind_rows(
  read_xpt(file.path(project_root, "data", "self-sleep-2011-12.xpt")),
  read_xpt(file.path(project_root, "data", "self-sleep-2013-14.xpt"))
) %>%
  transmute(ID = as.character(SEQN), self_reported_sleep_hours = SLD010H) %>%
  filter(!is.na(self_reported_sleep_hours), !(self_reported_sleep_hours %in% c(77, 99))) %>%
  left_join(participant_weights, by = "ID") %>%
  filter(!is.na(Exam_weight_recal), !is.na(PSU), !is.na(Stratum))

self_sleep_design <- svy_design(self_sleep)
self_sleep_est <- svymean(~self_reported_sleep_hours, self_sleep_design, na.rm = TRUE)
self_sleep_mean <- unname(coef(self_sleep_est)[1])
self_sleep_ci <- unname(confint(self_sleep_est)[1, ])
```

## Primary 1-Hour Cutoff And Participant Data

The final sleep/nonwear cutoff is selected from a fine-resolution objective
curve. For each candidate cutoff, the analysis compares survey-weighted
accelerometer-derived sleep/nonwear duration with the survey-weighted NHANES
self-reported sleep benchmark. The selected cutoff is the value that minimizes
the absolute difference between those two sleep-duration estimates. This
fine-resolution selected cutoff, not the coarse sensitivity-grid cutoff, is used
for the primary waking sedentary-time analysis.

```{r primary-cutoff}
objective_curve <- cutoff_objective_curve(primary_epoch, "swan", self_sleep_mean)
best_cutoff <- objective_curve %>%
  arrange(diff, cutoff) %>%
  slice(1)
selected_cutoff <- best_cutoff$cutoff[1]

analytic_df <- participant_from_epoch_cutoff(primary_epoch, selected_cutoff, "swan")
analytic_design <- svy_design(analytic_df)

write_csv(
  format_export_numbers(objective_curve),
  file.path(output_root, "cutoff", "objective_curve.csv")
)
write_csv(
  format_export_numbers(best_cutoff),
  file.path(output_root, "cutoff", "selected_cutoff.csv")
)
write_csv(
  format_export_numbers(analytic_df),
  file.path(output_root, "datasets", "analytic_participant_sedentary.csv")
)

objective_plot <- ggplot(objective_curve, aes(x = cutoff, y = diff)) +
  geom_line(color = "#2F5F98", linewidth = 1.1) +
  geom_point(data = best_cutoff, aes(x = cutoff, y = diff), color = "#8B1E2D", size = 3) +
  geom_vline(xintercept = selected_cutoff, color = "#8B1E2D", linewidth = 0.9, linetype = "dashed") +
  scale_x_continuous(breaks = seq(50, 100, 10), limits = c(50, 100)) +
  labs(
    x = "Sleep Cutoff %",
    y = "Absolute difference from self-report (h/day)"
  ) +
  theme_classic(base_size = 13)
save_final_figure(objective_plot, "objective_function_cutoff", width = 6.5, height = 4)
```

The fine-resolution objective curve selected a 1-hour SWaN sleep/nonwear cutoff
of `r format_decimal(selected_cutoff, 2)`%. At this cutoff, the
survey-weighted accelerometer-derived sleep/nonwear estimate was
`r format_decimal(best_cutoff$sleep_mean[1], 2)` hours/day, and the absolute
difference from the survey-weighted self-reported sleep benchmark was
`r format_decimal(best_cutoff$diff[1], 3)` hours/day.

## Main Estimates

Survey-weighted estimates are computed at the participant level. Each
participant contributes one usual daily waking sedentary-time estimate, one
waking-time estimate, and one sleep/nonwear-time estimate to the survey design.

```{r main-estimates}
get_main_results <- function(design_obj, self_sleep_design_obj) {
  bind_rows(
    survey_mean_row(design_obj, "est_sedentary_hours", "Waking sedentary time"),
    survey_mean_row(design_obj, "est_waking_hours", "Waking time"),
    survey_mean_row(design_obj, "est_sleep_nonwear_hours", "Sleep/nonwear time"),
    survey_mean_row(self_sleep_design_obj, "self_reported_sleep_hours", "Self-reported sleep time")
  )
}

main_results <- get_main_results(analytic_design, self_sleep_design)

cycle_main_results <- bind_rows(
  "Pooled" = main_results,
  "Cycle 2011-2012" = get_main_results(
    subset(analytic_design, Dataset == "2011-2012"),
    subset(self_sleep_design, Dataset == "2011-2012")
  ),
  "Cycle 2013-2014" = get_main_results(
    subset(analytic_design, Dataset == "2013-2014"),
    subset(self_sleep_design, Dataset == "2013-2014")
  ),
  .id = "Analysis group"
)

cycle_estimate_table <- cycle_main_results %>%
  filter(Measure != "Self-reported sleep time") %>%
  mutate(
    `Analysis group` = recode(
      `Analysis group`,
      "Pooled" = "Pooled",
      "Cycle 2011-2012" = "2011-2012 cycle",
      "Cycle 2013-2014" = "2013-2014 cycle"
    ),
    Measure = recode(
      Measure,
      "Waking sedentary time" = "Mean sedentary time",
      "Waking time" = "Waking time",
      "Sleep/nonwear time" = "Sleep/non-wear time"
    ),
    `95% CI` = paste0(
      format_decimal(`95% CI lower`, 2),
      "-",
      format_decimal(`95% CI upper`, 2)
    )
  ) %>%
  select(`Analysis group`, Measure, `Mean (h/day)`, SE, `95% CI`)

get_composition_results <- function(design_obj) {
  bind_rows(
    survey_mean_row(design_obj, "est_sedentary_hours", "Mean sedentary time"),
    survey_mean_row(
      design_obj,
      "I(24 - est_sedentary_hours - est_sleep_nonwear_hours)",
      "Non-sedentary time"
    ),
    survey_mean_row(design_obj, "est_sleep_nonwear_hours", "Sleep/non-wear time")
  )
}

composition_cycle_results <- bind_rows(
  "Pooled" = get_composition_results(analytic_design),
  "Cycle 2011-2012" = get_composition_results(subset(analytic_design, Dataset == "2011-2012")),
  "Cycle 2013-2014" = get_composition_results(subset(analytic_design, Dataset == "2013-2014")),
  .id = "Analysis group"
)

composition_cycle_table <- composition_cycle_results %>%
  mutate(
    `Analysis group` = recode(
      `Analysis group`,
      "Pooled" = "Pooled",
      "Cycle 2011-2012" = "2011-2012 cycle",
      "Cycle 2013-2014" = "2013-2014 cycle"
    ),
    `95% CI` = paste0(
      format_decimal(`95% CI lower`, 2),
      "-",
      format_decimal(`95% CI upper`, 2)
    )
  ) %>%
  select(`Analysis group`, Measure, `Mean (h/day)`, SE, `95% CI`)

write_csv(
  format_export_numbers(main_results),
  file.path(output_root, "results", "main_results.csv")
)
write_csv(
  format_export_numbers(cycle_main_results),
  file.path(output_root, "results", "main_results_by_cycle.csv")
)
write_csv(
  format_export_numbers(cycle_estimate_table),
  file.path(output_root, "results", "main_estimates_by_cycle_table.csv")
)
write_csv(
  format_export_numbers(composition_cycle_table),
  file.path(output_root, "results", "main_estimates_by_cycle_table_24h_composition.csv")
)

overall_sed <- svymean(~est_sedentary_hours, analytic_design, na.rm = TRUE)
overall_wake <- svymean(~est_waking_hours, analytic_design, na.rm = TRUE)
overall_sleep <- svymean(~est_sleep_nonwear_hours, analytic_design, na.rm = TRUE)

paper_kable(
  main_results,
  caption = "Primary SWaN 1-hour survey-weighted estimates."
)
```

```{r main-estimates-by-cycle-table}
paper_kable(
  cycle_estimate_table,
  caption = "Main survey-weighted estimates overall and by NHANES cycle.",
  align = c("l", "l", "r", "r", "r"),
  latex_size = "\\scriptsize"
)
```

```{r main-estimates-by-cycle-composition-table}
paper_kable(
  composition_cycle_table,
  caption = "Main survey-weighted estimates overall and by NHANES cycle with non-sedentary time.",
  align = c("l", "l", "r", "r", "r"),
  latex_size = "\\scriptsize"
)
```

Note. In the alternative table, non-sedentary time is calculated as 24 hours
minus sedentary time minus sleep/non-wear time, so the three estimates sum to a
24-hour day within each analysis group.

Note. Estimates are survey-weighted and expressed in hours per day. SE = standard
error; CI = confidence interval. Pooled estimates combine the 2011-2012 and
2013-2014 NHANES cycles using recalibrated 4-year examination weights.

```{r cycle-comparison}
cycle_design <- svy_design(
  analytic_df %>%
    mutate(Dataset = factor(Dataset, levels = c("2011-2012", "2013-2014")))
)

cycle_sed_estimates <- svyby(
  ~est_sedentary_hours,
  ~Dataset,
  cycle_design,
  svymean,
  na.rm = TRUE,
  vartype = "se"
) %>%
  as_tibble() %>%
  transmute(
    Dataset,
    `Mean sedentary time (hours/day)` = est_sedentary_hours,
    SE = se
  )

cycle_model <- svyglm(est_sedentary_hours ~ Dataset, design = cycle_design)
cycle_difference <- coef(cycle_model)["Dataset2013-2014"]
cycle_difference_se <- SE(cycle_model)["Dataset2013-2014"]
cycle_p_value <- as.numeric(summary(cycle_model)$coefficients["Dataset2013-2014", "Pr(>|t|)"])

cycle_comparison <- tibble(
  contrast = "2013-2014 minus 2011-2012",
  difference_hours = unname(cycle_difference),
  SE = unname(cycle_difference_se),
  p_value = cycle_p_value
)

cycle_2011 <- cycle_sed_estimates %>% filter(Dataset == "2011-2012") %>% slice(1)
cycle_2013 <- cycle_sed_estimates %>% filter(Dataset == "2013-2014") %>% slice(1)
cycle_2011_mean <- cycle_2011[["Mean sedentary time (hours/day)"]][1]
cycle_2011_se <- cycle_2011[["SE"]][1]
cycle_2013_mean <- cycle_2013[["Mean sedentary time (hours/day)"]][1]
cycle_2013_se <- cycle_2013[["SE"]][1]

write_csv(
  format_export_numbers(cycle_comparison),
  file.path(output_root, "results", "cycle_comparison.csv")
)

```

### Justification for Combining Survey Cycles

To assess whether the two NHANES cycles could be pooled, we compared
survey-weighted estimates of mean sedentary time between 2011-2012 and
2013-2014. The mean sedentary time was
`r format_decimal(cycle_2011_mean, 2)` hours/day
(SE = `r format_decimal(cycle_2011_se, 2)`) in 2011-2012 and
`r format_decimal(cycle_2013_mean, 2)` hours/day
(SE = `r format_decimal(cycle_2013_se, 2)`) in 2013-2014. The
survey-weighted regression estimate of the difference was
`r format_decimal(cycle_difference, 2)` hours/day
(p = `r format_decimal(cycle_p_value, 4)`). This difference is small relative to
the overall mean, so the two cycles were combined for subsequent analyses using
recalibrated 4-year examination weights.

### NHANES Analytic Approach

The primary estimand in this manuscript is the population mean of each
participant's usual daily waking sedentary time. The analysis therefore treats
the NHANES sample person, not the hour, as the survey-weighted analytic unit.
This matches NHANES guidance: sample weights are assigned to sample persons and
account for unequal probability of selection, nonresponse, and
post-stratification to U.S. population control totals.

The repeated accelerometer records were still used fully to derive the
participant-level exposure. For each candidate sleep cutoff, SWaN sleep/nonwear
percentages determined whether an epoch was treated as waking time, CHAP sitting
probabilities contributed fractional sedentary hours using the epoch duration,
daily totals were computed, and valid days were averaged within participant. The
person-level mean is therefore a summary of all available epoch- and daily-level
information for that participant.

What is intentionally removed before the main survey estimation is the repeated
record structure itself, because leaving epochs or days as separate weighted
records would give participants with more observed records or valid days more
influence than participants with fewer records. That would change the estimand
from a population mean across people to a record-weighted mean across observed
person-epochs or person-days.

### NHANES Sleep/Wake/Nonwear Algorithm

The NHANES comparison analysis used an open-source algorithm applied to the raw
80 Hz wrist accelerometer signal to classify each monitored period as wake
wear, sleep wear, non-wear, or unknown. The algorithm also assigned a confidence
score from 0.0 to 1.0 for each possible class. Most records received a
minute-level classification, but periods were left as unknown when the algorithm
could not assign a class with sufficient certainty.

The algorithm worked in three main steps. First, it summarized the raw
acceleration signal in short windows and used machine learning to classify each
30-second interval as wake wear, sleep wear, or non-wear. The features included
movement and orientation information, such as signal power, variability in
device orientation, and changes in orientation. The 30-second interval was used
because the sleep-validation data were scored against polysomnography in
30-second sleep/wake epochs.

This first step also addressed the accelerometer's idle sleep mode. Idle sleep
mode is a battery-saving feature that can remove useful low-movement signal
information. When idle sleep mode occurred, the algorithm replaced the missing
idle-mode signal with the 10 seconds of accelerometer data immediately before
idle sleep mode began. This copied segment represented the participant's
low-movement pattern just before the device entered idle mode. Although this is
only an approximation, validation testing suggested that it improved the
algorithm's ability to distinguish quiet waking behavior, sleep, and non-wear.

Second, the algorithm smoothed very short predicted periods. Wake-wear periods
shorter than 3 minutes, sleep-wear periods shorter than 10 minutes, and non-wear
periods shorter than 10 minutes were reconsidered using the classifications and
confidence scores from the periods immediately before and after them. If both
neighboring periods had the same class and their combined confidence was much
stronger than the confidence for the short middle period, the middle period was
changed to match its neighbors. If the neighboring periods had different
classes, the short middle period was changed to match the neighbor with the
higher confidence.

Third, the algorithm reviewed longer intervals using changes in device
orientation. This step allowed some longer periods to be reclassified between
sleep and non-wear. The purpose was to preserve sleep periods that included
occasional body reorientation, which is commonly observed when people change
position in bed during sleep.

## Table 1

```{r table-1}
df_days <- analytic_df %>%
  distinct(ID, Dataset, no_of_valid_days)

table_by_cycle <- df_days %>%
  count(Dataset, no_of_valid_days, name = "n") %>%
  group_by(Dataset) %>%
  mutate(
    percent = 100 * n / sum(n),
    value = paste0(comma(n), " (", format_decimal(percent, 1), ")")
  ) %>%
  ungroup() %>%
  select(no_of_valid_days, Dataset, value) %>%
  pivot_wider(names_from = Dataset, values_from = value)

table_total <- df_days %>%
  count(no_of_valid_days, name = "n") %>%
  mutate(
    percent = 100 * n / sum(n),
    `Total (2011-14)` = paste0(comma(n), " (", format_decimal(percent, 1), ")")
  ) %>%
  select(no_of_valid_days, `Total (2011-14)`)

sample_row <- df_days %>%
  group_by(Dataset) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(value = paste0(comma(n), " (", format_decimal(100, 1), ")")) %>%
  select(Dataset, value) %>%
  pivot_wider(names_from = Dataset, values_from = value) %>%
  mutate(
    no_of_valid_days = "Sample Analyzed (1-7)",
    `Total (2011-14)` = paste0(comma(nrow(df_days)), " (", format_decimal(100, 1), ")")
  )

table1 <- table_by_cycle %>%
  left_join(table_total, by = "no_of_valid_days") %>%
  arrange(no_of_valid_days) %>%
  mutate(no_of_valid_days = as.character(no_of_valid_days)) %>%
  bind_rows(sample_row) %>%
  select(
    `Full days` = no_of_valid_days,
    `2011-2012`,
    `2013-2014`,
    `Total (2011-14)`
  )

write_csv(table1, file.path(output_root, "tables", "table1_wear_day_distribution.csv"))

paper_kable(
  table1,
  caption = "Analytic sample and wear-day distribution by NHANES cycle (2011-2014)."
)
```

Notes: A full day is a 24-hour day represented by 24 hourly records; partial
initiation and return days are excluded.

## Epoch Sensitivity Table

```{r epoch-sensitivity}
analysis_epoch_specs <- tibble(
  epoch_level = c("10m", "30m", "1h", "90m", "120m", "150m", "180m"),
  minutes = c(10, 30, 60, 90, 120, 150, 180),
  label = c(
    "10-min epoch",
    "30-min epoch",
    "1-hour epoch",
    "1-hour 30-min epoch",
    "2-hour epoch",
    "2-hour 30-min epoch",
    "3-hour epoch"
  )
)

analysis_epoch_cache <- new.env(parent = emptyenv())

read_cached_epoch <- function(sleep_method, epoch_level) {
  cache_key <- paste(sleep_method, epoch_level, sep = "_")
  if (!exists(cache_key, envir = analysis_epoch_cache, inherits = FALSE)) {
    assign(cache_key, read_epoch(sleep_method, epoch_level), envir = analysis_epoch_cache)
  }
  get(cache_key, envir = analysis_epoch_cache, inherits = FALSE)
}

aggregate_analysis_epoch <- function(epoch_df, minutes, epoch_level) {
  percent_columns <- c(
    "percent_sedentary",
    "percent_waking_sedentary",
    "percent_sleep_nonwear",
    "percent_wake",
    "percent_missing_sleep"
  )

  weighted_percent <- function(x, duration) {
    if (all(is.na(x))) return(NA_real_)
    weighted.mean(x, duration, na.rm = TRUE)
  }

  epoch_df %>%
    mutate(
      Epoch = floor(epoch_start_minute / minutes) + 1L,
      source_duration_minutes = epoch_duration_minutes
    ) %>%
    group_by(ID, Dataset, Day, DayType, Epoch, sleep_method) %>%
    summarise(
      epoch_level = .env$epoch_level,
      epoch_start_minute = min(epoch_start_minute, na.rm = TRUE),
      epoch_duration_minutes = sum(source_duration_minutes, na.rm = TRUE),
      across(all_of(percent_columns), ~weighted_percent(.x, source_duration_minutes)),
      n_source_epochs = sum(n_source_epochs, na.rm = TRUE),
      expected_source_epochs = sum(expected_source_epochs, na.rm = TRUE),
      source_coverage = n_source_epochs / expected_source_epochs,
      .groups = "drop"
    )
}

read_analysis_epoch <- function(sleep_method, epoch_level) {
  if (epoch_level %in% c("10m", "30m", "1h")) {
    if (sleep_method == "swan" && epoch_level == "1h") return(primary_epoch)
    if (epoch_level %in% c("10m", "30m")) return(read_cached_epoch(sleep_method, epoch_level))
    return(read_epoch(sleep_method, epoch_level))
  }

  minutes <- analysis_epoch_specs %>%
    filter(epoch_level == .env$epoch_level) %>%
    pull(minutes)

  aggregate_analysis_epoch(read_cached_epoch(sleep_method, "30m"), minutes, epoch_level)
}

direct_state_row <- function(method_label, sleep_method, source_epoch_label) {
  person <- participant_from_direct_state(read_cached_epoch(sleep_method, "10m"), sleep_method)
  estimates <- estimate_three_means(person)
  tibble(
    Panel = method_label,
    Row = paste0("Direct states (", source_epoch_label, ")"),
    Cutoff = "-",
    `Mean sedentary hours` = estimates$mean_sedentary_hours,
    `SE sedentary hours` = estimates$se_sedentary_hours,
    `Mean waking hours` = estimates$mean_waking_hours,
    `SE waking hours` = estimates$se_waking_hours,
    `Mean sleep/nonwear hours` = estimates$mean_sleep_nonwear_hours,
    `SE sleep/nonwear hours` = estimates$se_sleep_nonwear_hours
  )
}

epoch_sensitivity_rows <- function(method_label, sleep_method) {
  rows <- vector("list", nrow(analysis_epoch_specs))

  for (i in seq_len(nrow(analysis_epoch_specs))) {
    epoch_level <- analysis_epoch_specs$epoch_level[[i]]
    label <- analysis_epoch_specs$label[[i]]
    epoch_df <- read_analysis_epoch(sleep_method, epoch_level)
    selected <- select_cutoff_for_epoch(epoch_df, sleep_method, self_sleep_mean)
    rows[[i]] <- tibble(
      Panel = method_label,
      Row = label,
      Cutoff = format_decimal(selected$cutoff, 2),
      `Mean sedentary hours` = selected$mean_sedentary_hours,
      `SE sedentary hours` = selected$se_sedentary_hours,
      `Mean waking hours` = selected$mean_waking_hours,
      `SE waking hours` = selected$se_waking_hours,
      `Mean sleep/nonwear hours` = selected$mean_sleep_nonwear_hours,
      `SE sleep/nonwear hours` = selected$se_sleep_nonwear_hours
    )

    rm(epoch_df, selected)
    invisible(gc())
  }

  bind_rows(rows)
}

sensitivity_table <- bind_rows(
  direct_state_row("SWaN sleep/wake/nonwear algorithm", "swan", "10-sec epoch"),
  epoch_sensitivity_rows("SWaN sleep/wake/nonwear algorithm", "swan"),
  direct_state_row("NHANES 1-minute sleep-state algorithm", "nhanes", "1-min epoch"),
  epoch_sensitivity_rows("NHANES 1-minute sleep-state algorithm", "nhanes")
)

write_csv(
  format_export_numbers(sensitivity_table),
  file.path(output_root, "tables", "epoch_sensitivity_table.csv")
)

paper_kable(
  sensitivity_table %>%
    transmute(
      Method = recode(
        Panel,
        "SWaN sleep/wake/nonwear algorithm" = "SWaN",
        "NHANES 1-minute sleep-state algorithm" = "NHANES"
      ),
      Row,
      Cutoff,
      Sedentary = format_mean_se(`Mean sedentary hours`, `SE sedentary hours`),
      Waking = format_mean_se(`Mean waking hours`, `SE waking hours`),
      `Sleep/nonwear` = format_mean_se(`Mean sleep/nonwear hours`, `SE sleep/nonwear hours`)
    ),
  caption = "Analysis epoch sensitivity analysis by sleep/wake detection algorithm.",
  latex_size = "\\scriptsize"
)
```

Notes: Direct states do not apply a sleep/nonwear cutoff. For SWaN, direct
states use the underlying SWaN wake, sleep, and non-wear classifications
combined with CHAP waking-sedentary time. For NHANES, direct states use the
official NHANES 1-minute sleep-state classifications combined with the same
CHAP posture summaries. Epoch rows first aggregate sleep/nonwear states to the
listed epoch length, then classify each whole epoch as waking when its
sleep/nonwear percentage is below the calibrated cutoff; epochs at or above the
cutoff are counted as sleep/non-wear.

## Cutoff Sensitivity Grid

This step does not select the final sleep/nonwear cutoff. It evaluates a coarse
grid of candidate cutoffs in 5-percentage-point increments to summarize how
survey-weighted sleep/nonwear duration and waking sedentary time change across
plausible decision rules. The final cutoff used in the primary analysis is the
fine-resolution value selected above.

```{r cutoff-sensitivity-grid}
sensitivity_cutoffs <- sort(unique(c(seq(50, 100, by = 5), selected_cutoff)))
sensitivity_results <- map_dfr(sensitivity_cutoffs, function(cut) {
  person <- if (isTRUE(all.equal(cut, selected_cutoff))) {
    analytic_df
  } else {
    participant_from_epoch_cutoff(primary_epoch, cut, "swan")
  }
  estimate_cutoff_curve_row(person, cut)
})

write_csv(
  format_export_numbers(sensitivity_results),
  file.path(output_root, "cutoff", "cutoff_sensitivity_grid.csv")
)
```

## Demographic Missingness

Demographic, anthropometric, and survey-design variables are taken from the
participant-level analysis datasets and reduced to one record per participant
before survey estimation. Age is grouped as `<18`, `18-34`, `35-44`, `45-64`,
and `>=65` years. Race/ethnicity is retained using the NHANES-coded categories
available in the analysis data. BMI is grouped as under/normal weight
(`BMI < 25`), overweight (`25 <= BMI < 30`), and obesity (`BMI >= 30`).
Participants with missing subgroup information are not recoded into another
category; missing values are excluded only from the relevant subgroup/domain
estimate.

These variables describe population heterogeneity in waking sedentary time, not
repeated epoch-level covariates. Subgroup estimates for age, gender,
race/ethnicity, and BMI are calculated from the participant-level survey design
object using domain analyses. This preserves the NHANES design information for
standard error estimation while estimating the mean for the population domain of
interest.

```{r demographic-missingness}
summarise_missingness <- function(df, variable, label, denominator) {
  values <- df[[variable]]
  missing_n <- sum(is.na(values))
  tibble(
    Variable = label,
    `Analytic sample N` = denominator,
    `Missing n` = missing_n,
    `Missing %` = 100 * missing_n / denominator,
    `Available n` = denominator - missing_n
  )
}

analytic_sample_n <- n_distinct(analytic_df$ID)
demographic_missingness <- bind_rows(
  summarise_missingness(analytic_df, "Age_group", "Age group", analytic_sample_n),
  summarise_missingness(analytic_df, "Gender", "Gender", analytic_sample_n),
  summarise_missingness(analytic_df, "Race", "Race/ethnicity", analytic_sample_n),
  summarise_missingness(analytic_df, "BMI_cat", "BMI category", analytic_sample_n),
  summarise_missingness(analytic_df, "Education", "Education", analytic_sample_n),
  summarise_missingness(analytic_df, "Marital_status", "Marital status", analytic_sample_n),
  summarise_missingness(analytic_df, "Exam_weight_recal", "Recalibrated examination weight", analytic_sample_n),
  summarise_missingness(analytic_df, "PSU", "Masked PSU", analytic_sample_n),
  summarise_missingness(analytic_df, "Stratum", "Masked stratum", analytic_sample_n)
) %>%
  arrange(Variable)

write_csv(
  format_export_numbers(demographic_missingness),
  file.path(output_root, "qc", "demographic_missingness.csv")
)

paper_kable(
  demographic_missingness,
  caption = "Missingness of demographic and survey design variables relative to the final analytic sample."
)
```

The final analytic sample included `r comma(analytic_sample_n)` participants
after applying accelerometer-data availability and survey-design requirements.
Missingness in demographic and survey-design variables is shown relative to this
analytic denominator so subgroup tables can be interpreted with the correct
available sample for each domain.

### Overview of Sedentary-Time Estimation

The analysis proceeded in the following steps:

1. **Start with wrist accelerometer data.** We used NHANES 2011-2012 and
2013-2014 wrist accelerometer files.

2. **Prepare the files for prediction models.** The raw files were organized by
participant and day and converted into the formats needed by the posture and
sleep/wear models upstream of this report.

3. **Identify sitting in 10-second epochs.** The CHAP/DeepPostures model
classified each 10-second epoch as sitting or not sitting.

4. **Identify sleep and nonwear.** SWaN classified accelerometer records as
wake wear, sleep, or nonwear; these predictions were summarized to the analysis
epochs used in this report.

5. **Summarize predictions for analysis.** Epoch-level predictions were
converted to epoch-level percentages for the primary analysis.

6. **Choose the sleep/nonwear cutoff.** Candidate cutoffs were evaluated by
comparing survey-weighted accelerometer-derived sleep/nonwear duration with
NHANES self-reported sleep duration.

7. **Keep only waking epochs for sedentary-time estimation.** Epochs below the
selected sleep/nonwear cutoff were treated as waking epochs. Epochs at or above
the cutoff were removed from waking sedentary-time estimation.

8. **Calculate sedentary hours for each day.** For each participant-day,
sedentary percentages were multiplied by epoch duration and added across waking
epochs.

9. **Average days within each participant.** Daily waking sedentary-time values
were averaged across each participant's valid wear days.

10. **Use the NHANES survey design.** Each participant's usual daily waking
sedentary time was analyzed using NHANES examination weights, masked variance
strata, and masked primary sampling units.

11. **Estimate national and subgroup values.** Survey-weighted means and
confidence intervals were estimated for the full population and for subgroups
defined by age, gender, race/ethnicity, and BMI category.

```{=latex}
\clearpage
```

## Figure 1

Workflow for deriving hourly sedentary and sleep/nonwear summaries. (A) Raw
80 Hz accelerometer signal. (B) CHAP sedentary predictions and SWaN
sleep/wake/nonwear classifications. (C) Hourly sedentary and sleep/nonwear
percentages used to define waking sedentary time under candidate sleep cutoffs.

```{r figure-1, fig.width=7, fig.height=5.8, out.width="74%"}
workflow_env <- new.env(parent = globalenv())
final_workflow_png <- file.path(figure_root, "fig1_workflow.png")
final_workflow_pdf <- file.path(figure_root, "fig1_workflow.pdf")

local({
  old_workflow_dir <- Sys.getenv("WORKFLOW_FINAL_FIGURE_DIR", unset = NA_character_)
  on.exit({
    if (is.na(old_workflow_dir)) {
      Sys.unsetenv("WORKFLOW_FINAL_FIGURE_DIR")
    } else {
      Sys.setenv(WORKFLOW_FINAL_FIGURE_DIR = old_workflow_dir)
    }
  })

  Sys.setenv(WORKFLOW_FINAL_FIGURE_DIR = figure_root)
  invisible(capture.output(
    source(file.path(project_root, "R", "workflow_figure.R"), local = workflow_env)
  ))
})

if (!file.exists(final_workflow_png) || !file.exists(final_workflow_pdf)) {
  stop("The preprocessing workflow script did not produce the expected final Figure 1 files.")
}

knitr::include_graphics(
  normalizePath(final_workflow_png, winslash = "/", mustWork = TRUE),
  rel_path = FALSE
)
```

## Table 2

This table places the current wrist-derived estimate beside prior NHANES
sedentary-time estimates that used different devices, time periods, populations,
or self-report instruments.

```{r table-2}
table2 <- tibble::tribble(
  ~Paper, ~`Dataset & Population`, ~Method, ~`Estimate & Notes`,
  "This study",
  paste0(
    "NHANES 2011-2014, wrist-worn 80 Hz accelerometer data; ",
    comma(n_distinct(analytic_df$ID)),
    " participants, aged 3+ years."
  ),
  "CHAP wrist model combined with SWaN.",
  paste0(
    "Sedentary time: ",
    format_decimal(coef(overall_sed)[1], 1),
    " hours/day; ",
    format_decimal(100 * coef(overall_sed)[1] / coef(overall_wake)[1], 1),
    "% of non-sleep time. Mean waking time was ",
    format_decimal(coef(overall_wake)[1], 1),
    " hours/day after excluding sleep."
  ),
  "Matthews et al. (2008)",
  "NHANES 2003-2004; hip-worn ActiGraph data; 6,329 participants, aged 6+ years.",
  "Intensity threshold of <100 counts/min.",
  "Sedentary time: 7.7 hours/day; 54.9% of monitored time. Mean monitor wear time was 13.9 hours/day.",
  "Yang et al. (2019)",
  "NHANES 2001-2016; 51,896 children and adolescents.",
  "Self-reported total sitting time from questionnaire.",
  "Sedentary time: 6.4 hours/day in adults; 8.2 hours/day in adolescents. Sitting time increased in both groups from 2001 onward.",
  "Ussery et al. (2018)",
  "NHANES 2015-2016; 5,923 U.S. adults aged 18+ years.",
  "Self-reported sitting time on a typical day.",
  "~6.4 hours/day. Excludes sleeping time; 25.7% of adults reported sitting for more than 8 hours per day."
)

write_csv(table2, file.path(output_root, "tables", "table2_study_comparison.csv"))

paper_kable(
  table2,
  caption = "Comparison of sedentary time estimates across studies by measurement method, datasets, and population.",
  align = c("p{0.16\\linewidth}", "p{0.29\\linewidth}", "p{0.20\\linewidth}", "p{0.27\\linewidth}"),
  latex_size = "\\scriptsize"
)
```

## Figure 2

Survey-weighted hourly patterns of wrist-derived waking sedentary behavior are
shown by day type after excluding records with sleep/nonwear greater than or
equal to the selected `r format_decimal(selected_cutoff, 2)`% cutoff. Lines
represent hourly means and shaded bands indicate 95% confidence intervals.

```{r figure-2, fig.width=7.5, fig.height=4.8}
hourly_df <- primary_epoch %>%
  filter(percent_sleep_nonwear < selected_cutoff) %>%
  mutate(
    Hour = Epoch,
    DayType = factor(DayType, levels = c("Weekend", "Weekday"))
  ) %>%
  left_join(participant_weights, by = c("ID", "Dataset")) %>%
  filter(!is.na(Exam_weight_recal), !is.na(PSU), !is.na(Stratum))

hourly_design <- svy_design(hourly_df)

hourly_est <- svyby(
  ~percent_sedentary,
  ~DayType + Hour,
  hourly_design,
  svymean,
  vartype = "ci",
  na.rm = TRUE
) %>%
  as_tibble()

daytype_colors <- c("Weekend" = "#0072B2", "Weekday" = "#D55E00")

hourly_plot <- ggplot(hourly_est, aes(x = Hour, y = percent_sedentary, group = DayType)) +
  geom_ribbon(aes(ymin = ci_l, ymax = ci_u, fill = DayType), alpha = 0.28, color = NA) +
  geom_line(aes(color = DayType, linetype = DayType), linewidth = 1.25) +
  geom_point(aes(color = DayType, shape = DayType), size = 2.7, stroke = 1.05, fill = "white") +
  scale_color_manual(values = daytype_colors) +
  scale_fill_manual(values = daytype_colors) +
  scale_linetype_manual(values = c("Weekend" = "solid", "Weekday" = "dashed")) +
  scale_shape_manual(values = c("Weekend" = 16, "Weekday" = 1)) +
  scale_x_continuous(breaks = seq(0, 24, 4), minor_breaks = seq(1, 24, 1), limits = c(1, 24)) +
  scale_y_continuous(labels = percent_format(scale = 1), limits = c(50, 100), breaks = seq(50, 100, 10)) +
  labs(x = "Hour of day", y = "Waking sedentary (%)", color = NULL, fill = NULL, linetype = NULL, shape = NULL) +
  guides(fill = "none") +
  theme_bw(base_size = 13) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    panel.grid.major.x = element_line(color = "grey90", linewidth = 0.3),
    panel.grid.minor.x = element_line(color = "grey95", linewidth = 0.2),
    panel.grid.minor.y = element_blank(),
    legend.position = "bottom"
  )

save_final_figure(hourly_plot, "figure2_hourly_waking_sedentary", width = 7.5, height = 4.8)
ggsave(file.path(figure_root, "hourly_waking_sedentary.png"), hourly_plot, width = 7.5, height = 4.8, dpi = 300, bg = "white")
write_csv(
  format_export_numbers(hourly_est),
  file.path(output_root, "results", "hourly_waking_sedentary.csv")
)
hourly_plot
```

## Figure 3

Distribution of participants by mean daily waking sedentary time and
corresponding 24-hour behavioral composition. Panel A shows the proportion and
unweighted number of participants in each sedentary-time category. Panel B shows
the mean daily allocation to sedentary time, sleep/nonwear, and non-sedentary
waking time within each category.

```{r figure-3, fig.width=8, fig.height=3.8, out.width="100%"}
df_plot <- analytic_df %>%
  distinct(ID, .keep_all = TRUE) %>%
  mutate(
    sed_bin = cut(
      est_sedentary_hours,
      breaks = c(0, 4, seq(6, 16, by = 2), Inf),
      labels = c("0-4", "4-6", "6-8", "8-10", "10-12", "12-14", "14-16", "16+"),
      include.lowest = TRUE,
      right = FALSE
    )
  ) %>%
  filter(!is.na(sed_bin))

plot_df <- df_plot %>%
  group_by(sed_bin) %>%
  summarise(
    n = n_distinct(ID),
    prop = n / n_distinct(df_plot$ID),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("n=", comma(n), "\n", percent(prop, accuracy = 0.1)))

figure3_theme <- theme_bw(base_size = 16) +
  theme(
    axis.title = element_text(face = "bold", size = 15),
    axis.text = element_text(color = "black", face = "bold", size = 13),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", linewidth = 0.95),
    axis.ticks = element_line(color = "black", linewidth = 0.75),
    axis.ticks.length = grid::unit(0.12, "in")
  )

left_plot <- ggplot(plot_df, aes(x = sed_bin, y = prop)) +
  geom_col(width = 0.5, fill = "#4E79A7", color = "black", linewidth = 0.5) +
  geom_text(aes(label = label), vjust = -0.35, size = 4.1, lineheight = 0.86, fontface = "bold") +
  scale_x_discrete(expand = expansion(add = 0.7)) +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.24))) +
  labs(x = "Sedentary time (h/day)", y = "Participants (%)", tag = "A") +
  coord_cartesian(clip = "off") +
  figure3_theme +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1, vjust = 1),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey86", linewidth = 0.45)
  )

right_df <- df_plot %>%
  group_by(sed_bin) %>%
  summarise(
    Sedentary = mean(est_sedentary_hours, na.rm = TRUE),
    `Sleep/nonwear` = mean(est_sleep_nonwear_hours, na.rm = TRUE),
    `Non-sedentary` = mean(24 - est_sedentary_hours - est_sleep_nonwear_hours, na.rm = TRUE),
    .groups = "drop"
  )

right_df_long <- right_df %>%
  pivot_longer(
    cols = c(`Sleep/nonwear`, Sedentary, `Non-sedentary`),
    names_to = "behavior",
    values_to = "hours"
  ) %>%
  mutate(behavior = factor(behavior, levels = c("Sleep/nonwear", "Sedentary", "Non-sedentary")))

right_label_df <- right_df %>%
  pivot_longer(
    cols = c(`Non-sedentary`, Sedentary, `Sleep/nonwear`),
    names_to = "behavior",
    values_to = "hours"
  ) %>%
  group_by(sed_bin) %>%
  mutate(
    x_mid = cumsum(hours) - hours / 2,
    percent_label = ifelse(hours >= 1.2, percent(hours / 24, accuracy = 1), ""),
    label_color = ifelse(behavior %in% c("Non-sedentary", "Sedentary"), "white", "black")
  ) %>%
  ungroup()

right_plot <- ggplot(right_df_long, aes(x = hours, y = sed_bin, fill = behavior)) +
  geom_col(width = 0.72, color = "white", linewidth = 0.65) +
  geom_text(
    data = right_label_df,
    aes(x = x_mid, y = sed_bin, label = percent_label, color = label_color),
    inherit.aes = FALSE,
    size = 4.25,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c("Sleep/nonwear" = "#9ECAE1", "Sedentary" = "#4E79A7", "Non-sedentary" = "#59A14F")) +
  scale_color_identity(guide = "none") +
  guides(fill = guide_legend(reverse = TRUE)) +
  scale_x_continuous(breaks = seq(0, 24, by = 4), expand = expansion(mult = c(0, 0.02))) +
  coord_cartesian(xlim = c(0, 24), clip = "off") +
  labs(x = "Hours/day", y = "Sedentary time (h/day)", fill = NULL, tag = "B") +
  figure3_theme +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "grey86", linewidth = 0.45),
    legend.position = "bottom",
    legend.text = element_text(size = 13, face = "bold")
  )

figure3 <- left_plot + right_plot +
  plot_layout(widths = c(1.55, 1.35), guides = "collect") &
  theme(
    legend.position = "bottom",
    plot.tag = element_text(face = "bold", size = 22),
    plot.tag.position = c(0.02, 0.98)
  )

save_final_figure(figure3, "sedentary_distribution_with_24h_composition", width = 12.8, height = 5.8)
ggsave(file.path(figure_root, "sedentary_distribution_composition.png"), figure3, width = 12.8, height = 5.8, dpi = 300, bg = "white")
knitr::include_graphics(
  normalizePath(file.path(figure_root, "sedentary_distribution_with_24h_composition.png"), winslash = "/", mustWork = TRUE),
  rel_path = FALSE
)
```

## Figure 3 Alternative

This version keeps the same 24-hour composition but labels the sleep/nonwear
component as Sleep in the legend.

```{r figure-3-sleep-label, fig.width=8, fig.height=3.8, out.width="100%"}
right_df_sleep_label <- right_df %>%
  rename(Sleep = `Sleep/nonwear`)

right_df_long_sleep_label <- right_df_sleep_label %>%
  pivot_longer(
    cols = c(Sleep, Sedentary, `Non-sedentary`),
    names_to = "behavior",
    values_to = "hours"
  ) %>%
  mutate(behavior = factor(behavior, levels = c("Sleep", "Sedentary", "Non-sedentary")))

right_label_df_sleep_label <- right_df_sleep_label %>%
  pivot_longer(
    cols = c(`Non-sedentary`, Sedentary, Sleep),
    names_to = "behavior",
    values_to = "hours"
  ) %>%
  group_by(sed_bin) %>%
  mutate(
    x_mid = cumsum(hours) - hours / 2,
    percent_label = ifelse(hours >= 1.2, percent(hours / 24, accuracy = 1), ""),
    label_color = ifelse(behavior %in% c("Non-sedentary", "Sedentary"), "white", "black")
  ) %>%
  ungroup()

right_plot_sleep_label <- ggplot(right_df_long_sleep_label, aes(x = hours, y = sed_bin, fill = behavior)) +
  geom_col(width = 0.72, color = "white", linewidth = 0.65) +
  geom_text(
    data = right_label_df_sleep_label,
    aes(x = x_mid, y = sed_bin, label = percent_label, color = label_color),
    inherit.aes = FALSE,
    size = 4.25,
    fontface = "bold"
  ) +
  scale_fill_manual(values = c("Sleep" = "#9ECAE1", "Sedentary" = "#4E79A7", "Non-sedentary" = "#59A14F")) +
  scale_color_identity(guide = "none") +
  guides(fill = guide_legend(reverse = TRUE)) +
  scale_x_continuous(breaks = seq(0, 24, by = 4), expand = expansion(mult = c(0, 0.02))) +
  coord_cartesian(xlim = c(0, 24), clip = "off") +
  labs(x = "Hours/day", y = "Sedentary time (h/day)", fill = NULL, tag = "B") +
  figure3_theme +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "grey86", linewidth = 0.45),
    legend.position = "bottom",
    legend.text = element_text(size = 13, face = "bold")
  )

figure3_sleep_label <- left_plot + right_plot_sleep_label +
  plot_layout(widths = c(1.55, 1.35), guides = "collect") &
  theme(
    legend.position = "bottom",
    plot.tag = element_text(face = "bold", size = 22),
    plot.tag.position = c(0.02, 0.98)
  )

save_final_figure(figure3_sleep_label, "sedentary_distribution_with_24h_composition_sleep_label", width = 12.8, height = 5.8)
knitr::include_graphics(
  normalizePath(file.path(figure_root, "sedentary_distribution_with_24h_composition_sleep_label.png"), winslash = "/", mustWork = TRUE),
  rel_path = FALSE
)
```

## Table 3

```{r table-3}
age_levels <- c("<18", "18-34", "35-44", "45-64", ">=65")
denominator <- n_distinct(analytic_df$ID)

make_age_gender_row <- function(age_group) {
  data_age <- if (age_group == "All ages") analytic_df else filter(analytic_df, Age_group == age_group)
  des_age <- if (age_group == "All ages") analytic_design else subset(analytic_design, Age_group == age_group)

  tibble(
    `Age group` = age_group,
    `Male n (%)` = format_n_pct(filter(data_age, Gender == "Male"), denominator),
    `Male mean (95% CI)` = safe_format_mean_ci(subset(des_age, Gender == "Male")),
    `Female n (%)` = format_n_pct(filter(data_age, Gender == "Female"), denominator),
    `Female mean (95% CI)` = safe_format_mean_ci(subset(des_age, Gender == "Female"))
  )
}

table3 <- map_dfr(c("All ages", age_levels), make_age_gender_row)

write_csv(table3, file.path(output_root, "tables", "table3_age_gender_sedentary.csv"))

paper_kable(
  table3,
  caption = "Weighted sedentary behavior by gender and age, United States, 2011-2014.",
  align = c("l", "r", "r", "r", "r"),
  latex_size = "\\scriptsize"
)
```

## Table 4

```{r table-4}
race_levels <- c("Mexican American", "NH Asian", "NH Black", "NH White", "Other Hispanic", "Other/Multi")
bmi_levels <- c("under/normal weight", "overweight", "obesity")

make_race_bmi_row <- function(race) {
  get_cell <- function(bmi_cat) {
    if (race == "All") {
      safe_format_mean_ci(subset(analytic_design, BMI_cat == bmi_cat))
    } else {
      safe_format_mean_ci(subset(analytic_design, Race == race & BMI_cat == bmi_cat))
    }
  }

  tibble(
    Race = race,
    `Underweight/normal` = get_cell("under/normal weight"),
    Overweight = get_cell("overweight"),
    Obesity = get_cell("obesity")
  )
}

table4 <- map_dfr(c("All", race_levels), make_race_bmi_row)

write_csv(table4, file.path(output_root, "tables", "table4_race_bmi_sedentary.csv"))

paper_kable(
  table4,
  caption = "Weighted mean sedentary time by race/ethnicity and BMI category, NHANES 2011-2014.",
  latex_size = "\\scriptsize"
)
```

## Table 5

```{r table-5}
make_marginal_subgroup_rows <- function(variable, subgroup_label, levels) {
  map_dfr(
    levels,
    function(level) {
      subgroup_data <- analytic_df %>% filter(.data[[variable]] == level)
      subgroup_design <- subset(analytic_design, analytic_design$variables[[variable]] == level)

      tibble(
        Subgroup = subgroup_label,
        Category = level,
        `N (%)` = format_n_pct(subgroup_data, denominator),
        `Mean sedentary time (95% CI), hours/day` = safe_format_mean_ci(subgroup_design)
      )
    }
  )
}

table5 <- bind_rows(
  make_marginal_subgroup_rows("Age_group", "Age group", age_levels),
  make_marginal_subgroup_rows("Gender", "Gender", c("Male", "Female")),
  make_marginal_subgroup_rows("Race", "Race/ethnicity", race_levels),
  make_marginal_subgroup_rows("BMI_cat", "BMI category", bmi_levels)
)

table5_render <- table5 %>%
  group_by(Subgroup) %>%
  mutate(Subgroup = if_else(row_number() == 1, Subgroup, "")) %>%
  ungroup()

write_csv(table5, file.path(output_root, "tables", "table5_marginal_subgroup_sedentary.csv"))

paper_kable(
  table5_render,
  caption = "Survey-weighted sedentary time by demographic subgroup, NHANES 2011-2014.",
  latex_size = "\\scriptsize"
)
```

## Sleep Distribution Diagnostic

```{r sleep-distribution-diagnostic, fig.width=6.5, fig.height=3.8, out.width="86%"}
sleep_distribution <- bind_rows(
  self_sleep %>%
    transmute(method = "Self-reported sleep", sleep_hours = self_reported_sleep_hours),
  analytic_df %>%
    transmute(method = "Accelerometer sleep/nonwear", sleep_hours = est_sleep_nonwear_hours)
) %>%
  mutate(
    method = factor(method, levels = c("Self-reported sleep", "Accelerometer sleep/nonwear")),
    sleep_bin = cut(
      sleep_hours,
      breaks = c(0, 4, 6, 8, 10, 12, 14, Inf),
      labels = c("0-4", "4-6", "6-8", "8-10", "10-12", "12-14", "14+"),
      include.lowest = TRUE,
      right = FALSE
    )
  ) %>%
  filter(!is.na(sleep_bin)) %>%
  count(method, sleep_bin, name = "n") %>%
  complete(method, sleep_bin, fill = list(n = 0L)) %>%
  group_by(method) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  ungroup() %>%
  mutate(label = if_else(n > 0L, comma(n), ""))

sleep_bar_position <- position_dodge(width = 0.75)

sleep_distribution_plot <- ggplot(sleep_distribution, aes(x = sleep_bin, y = percent, fill = method)) +
  geom_col(position = sleep_bar_position, width = 0.68, color = "white") +
  geom_text(
    aes(label = label),
    position = sleep_bar_position,
    vjust = -0.25,
    size = 3.1,
    color = "black"
  ) +
  scale_y_continuous(labels = percent_format(scale = 1), expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(
    values = c("Self-reported sleep" = "#8B1E2D", "Accelerometer sleep/nonwear" = "#2F5F98"),
    labels = c("Self-reported sleep", "Accelerometer sleep")
  ) +
  labs(x = "Sleep duration (h/day)", y = "Participants (%)", fill = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom")

save_final_figure(sleep_distribution_plot, "sleep_distribution_comparison", width = 8, height = 4.5)
knitr::include_graphics(
  normalizePath(file.path(figure_root, "sleep_distribution_comparison.png"), winslash = "/", mustWork = TRUE),
  rel_path = FALSE
)
```

## Figure 4

Sleep-cutoff calibration and sedentary-time sensitivity. Survey-weighted mean
accelerometer-derived sleep/nonwear duration and waking sedentary time are shown
across candidate sleep-percentage cutoffs. Shaded bands indicate 95% confidence
intervals. The vertical dashed line marks the selected
`r format_decimal(selected_cutoff, 2)`% cutoff, calibrated to the self-reported
sleep-duration benchmark.

```{r figure-4, fig.width=7, fig.height=5.6, out.width="84%"}
figure4_df <- sensitivity_results %>%
  mutate(
    sleep_ci_l = sleep_mean - qnorm(0.975) * sleep_se,
    sleep_ci_u = sleep_mean + qnorm(0.975) * sleep_se,
    sed_ci_l = sed_mean - qnorm(0.975) * sed_se,
    sed_ci_u = sed_mean + qnorm(0.975) * sed_se
  )

selected_curve_row <- figure4_df %>%
  filter(cutoff == selected_cutoff) %>%
  slice(1)

fig4_theme <- theme_classic(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    plot.tag = element_text(face = "bold", size = 18),
    axis.title = element_text(face = "bold", size = 14),
    axis.text = element_text(color = "black"),
    axis.line = element_line(linewidth = 0.9, color = "black"),
    axis.ticks = element_line(linewidth = 0.9, color = "black"),
    legend.position = "none"
  )

sleep_panel <- ggplot(figure4_df, aes(x = cutoff, y = sleep_mean)) +
  annotate("rect", xmin = selected_cutoff - 1.25, xmax = selected_cutoff + 1.25, ymin = -Inf, ymax = Inf, fill = "#8B1E2D", alpha = 0.08) +
  geom_ribbon(aes(ymin = sleep_ci_l, ymax = sleep_ci_u), fill = "#D9E8F5", alpha = 0.9) +
  geom_line(color = "#2F5F98", linewidth = 1.25) +
  geom_point(color = "#2F5F98", fill = "white", shape = 21, stroke = 1.1, size = 2.8) +
  geom_hline(yintercept = self_sleep_mean, color = "#8B1E2D", linewidth = 0.9, linetype = "dashed") +
  geom_vline(xintercept = selected_cutoff, color = "#8B1E2D", linewidth = 0.9, linetype = "dashed") +
  annotate(
    "text",
    x = selected_cutoff,
    y = selected_curve_row$sleep_mean,
    label = paste0(format_decimal(selected_cutoff, 2), "%"),
    hjust = 1.15,
    vjust = 1.5,
    color = "#8B1E2D",
    fontface = "bold",
    size = 4
  ) +
  scale_x_continuous(breaks = seq(50, 100, 10), limits = c(50, 100)) +
  labs(tag = "A", title = "Sleep calibration", x = "Sleep Cutoff %", y = "Sleep (h/day)") +
  fig4_theme

sed_panel <- ggplot(figure4_df, aes(x = cutoff, y = sed_mean)) +
  annotate("rect", xmin = selected_cutoff - 1.25, xmax = selected_cutoff + 1.25, ymin = -Inf, ymax = Inf, fill = "#8B1E2D", alpha = 0.08) +
  geom_ribbon(aes(ymin = sed_ci_l, ymax = sed_ci_u), fill = "#E8DFF4", alpha = 0.9) +
  geom_line(color = "#7B3FC6", linewidth = 1.25) +
  geom_point(color = "#7B3FC6", fill = "white", shape = 21, stroke = 1.1, size = 2.8) +
  geom_vline(xintercept = selected_cutoff, color = "#8B1E2D", linewidth = 0.9, linetype = "dashed") +
  annotate(
    "text",
    x = selected_cutoff,
    y = selected_curve_row$sed_mean,
    label = paste0(format_decimal(selected_curve_row$sed_mean, 2), " h/day"),
    hjust = 1.1,
    vjust = -0.6,
    color = "#7B3FC6",
    fontface = "bold",
    size = 4
  ) +
  scale_x_continuous(breaks = seq(50, 100, 10), limits = c(50, 100)) +
  labs(tag = "B", title = "Sedentary-time sensitivity", x = "Sleep Cutoff %", y = "Sedentary (h/day)") +
  fig4_theme

figure4 <- sleep_panel / sed_panel + plot_layout(heights = c(1, 1))

save_final_figure(figure4, "cutoff_sensi", width = 9.5, height = 7.6)
knitr::include_graphics(
  normalizePath(file.path(figure_root, "cutoff_sensi.png"), winslash = "/", mustWork = TRUE),
  rel_path = FALSE
)
```

## Output Index

```{r output-index}
output_index <- tibble::tribble(
  ~Artifact, ~Path,
  "Main results", file.path(output_root, "results", "main_results.csv"),
  "Main results by cycle", file.path(output_root, "results", "main_results_by_cycle.csv"),
  "Main estimates by cycle table", file.path(output_root, "results", "main_estimates_by_cycle_table.csv"),
  "Main estimates by cycle 24-hour composition table", file.path(output_root, "results", "main_estimates_by_cycle_table_24h_composition.csv"),
  "Cycle comparison", file.path(output_root, "results", "cycle_comparison.csv"),
  "Table 1", file.path(output_root, "tables", "table1_wear_day_distribution.csv"),
  "Table 2", file.path(output_root, "tables", "table2_study_comparison.csv"),
  "Table 3", file.path(output_root, "tables", "table3_age_gender_sedentary.csv"),
  "Table 4", file.path(output_root, "tables", "table4_race_bmi_sedentary.csv"),
  "Table 5", file.path(output_root, "tables", "table5_marginal_subgroup_sedentary.csv"),
  "Epoch sensitivity table", file.path(output_root, "tables", "epoch_sensitivity_table.csv"),
  "Cutoff sensitivity grid", file.path(output_root, "cutoff", "cutoff_sensitivity_grid.csv"),
  "Objective curve", file.path(output_root, "cutoff", "objective_curve.csv"),
  "Selected cutoff", file.path(output_root, "cutoff", "selected_cutoff.csv"),
  "Demographic missingness", file.path(output_root, "qc", "demographic_missingness.csv"),
  "Analytic participant dataset", file.path(output_root, "datasets", "analytic_participant_sedentary.csv"),
  "Figure 1", file.path(figure_root, "fig1_workflow.png"),
  "Figure 2", file.path(figure_root, "figure2_hourly_waking_sedentary.png"),
  "Figure 3", file.path(figure_root, "sedentary_distribution_with_24h_composition.png"),
  "Figure 3 alternative", file.path(figure_root, "sedentary_distribution_with_24h_composition_sleep_label.png"),
  "Figure 4", file.path(figure_root, "cutoff_sensi.png"),
  "Objective diagnostic", file.path(figure_root, "objective_function_cutoff.png"),
  "Sleep distribution diagnostic", file.path(figure_root, "sleep_distribution_comparison.png")
) %>%
  mutate(
    Path = sub(
      paste0(normalizePath(project_root, winslash = "/", mustWork = TRUE), "/"),
      "",
      normalizePath(Path, winslash = "/", mustWork = FALSE),
      fixed = TRUE
    )
  )

paper_kable(
  output_index,
  caption = "Final analysis outputs.",
  align = c("p{0.28\\linewidth}", "p{0.66\\linewidth}"),
  latex_size = "\\scriptsize"
)
```
